---
title: "Multilevel Determinants of Player Market Value in Elite European Football"
format:
  pdf:
    documentclass: article
    keep-tex: true
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
    fontsize: 12pts
    include-in-header:
      text: |
        \usepackage{amsmath,amssymb}
        \usepackage{booktabs}
        \usepackage{threeparttable}
        \usepackage{setspace}
        \doublespacing
author:
  - name: Kevin Linares
    affiliations:
      - name: University of Maryland
  - name: Jianing Zou
    affiliations:
      - name: University of Maryland
  - name: Namit Shrivastava
    affiliations:
      - name: University of Maryland
  - name: Sagnik Chakravarti
    affiliations:
      - name: University of Maryland
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
#| cache: false

# Loading required packages
library(arrow)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(performance)
library(knitr)
library(kableExtra)
library(jtools)
library(ggeffects)
library(patchwork)
library(scales)
library(viridis)
library(ggdist)
library(arm)
library(tidyverse)

# Setting options
options(scipen = 999)
theme_set(theme_minimal(base_size = 12))

# Setting repository path for macOS
#repo_path <- "/Users/namomac/Desktop/Moneyball_FC/inputs/"
repo_path <- "D:/repos/Moneyball_FC/inputs/"

# Reading data
#dat <- read_parquet(paste0(repo_path, "player_df.parquet"))
dat <- read_parquet(str_c(repo_path, "player_df.parquet")) 

# source file
source("D:/repos/Moneyball_FC/Project_Paper_files/source_code.r" )
```

```{r data-prep}
# Computing sample statistics
n_obs <- nrow(clean_player_df)
n_players <- n_distinct(clean_player_df$player_id)
n_clubs <- n_distinct(clean_player_df$club_id)
n_years <- n_distinct(clean_player_df$season_year)
mean_age <- round(mean(dat$Age, na.rm = TRUE), 1)
sd_age <- round(sd(dat$Age, na.rm = TRUE), 1)

# Counting players who transferred between clubs
transfers <- clean_player_df |>
  distinct(player_id, club_id) |>
  count(player_id) |>
  filter(n > 1)

n_transferred <- nrow(transfers)
pct_transferred <- round(100 * n_transferred / n_players, 1)
```

```{r model-fitting}
# Model_3.2
final_model <- model_player
vc <- as.data.frame(VarCorr(final_model))

```

# Introduction

The European football transfer market has evolved from a subjective scouting exercise into a global financial industry exceeding €7 billion annually. For this study, we operationalize market value using crowd-sourced estimates from Transfermarkt, a metric widely validated in sports economics as a robust proxy for actual transfer fees (Herm et al., 2014). While foundational research identifies performance metrics like goals as primary drivers (Frick, 2007), more recent scholarship highlights a "Superstar Effect" where non-performance factors like brand visibility magnify valuation (Franck & Nüesch, 2012; Rosen, 1981). However, the market has recently bifurcated into a pre-2020 hyper-inflationary era, termed the "Neymar Effect" (Poli et al., 2017), and a post-2020 pandemic correction (Drewes et al., 2021). This trajectory was violently interrupted by the COVID-19 pandemic, which caused an estimated €8.7 billion revenue shortfall across European football (Deloitte, 2021). Existing models typically treat players as static units nested within single clubs, ignoring the dynamic reality where players transfer between teams and accumulate value from both their portable human capital and the institutional prestige of their current employer. Neglecting this cross-classified structure likely misestimates the true drivers of market value in this volatile era.\
\
To address these limitations, this study applies a cross-classified multilevel model to a decade of transfer data (2013–2024), centering the analysis at the 2020 pandemic baseline to quantify the impact of this structural break. By disentangling individual reputation from club context, we investigate four critical dynamics: the non-linear "Boom and Bust" market trajectory (RQ1)[^1], the biological constraints of the player career arc (RQ2)[^2], the valuation signal of "Squad Status" via relative playing time (RQ3)[^3], and the existence of contextual premiums for club diversity or discounts for domestic talent in a globalized labor market (RQ4[^4]-RQ5[^5]).

[^1]: How has the transfer market evolved relative to the 2020 baseline? Specifically, does the market exhibit a non-linear "Boom and Bust" cycle that linear models miss?

[^2]: How does market valuation evolve across a player's biological career, and at what specific age does this value peak?

[^3]: To what degree does "Squad Status"—defined by relative playing time—predict market value independent of raw performance output?

[^4]: How is club-level international diversity associated with player valuations in a globalized labor market?

[^5]: How does the market valuation of domestic players compare to that of imported talent when controlling for performance and club context?

# Method

## Sample and Variables

We obtained data from Transfermarkt and UEFA official rankings, both publicly available sources. We used purposive sampling to select the top 21 European clubs based on 10-year averaged UEFA coefficient rankings[^6]. We then identified all players who played a full season in any of these clubs from 2013 through 2024 resulting in a sample of `r format(n_obs, big.mark = ",")` player-season observations from `r format(n_players, big.mark = ",")` unique players across `r n_clubs` clubs over `r n_years` seasons. After listwise deletion of cases with missing citizenship data (approximately 2% loss), players averaged `r mean_age` years of age (*SD* = `r sd_age`). Critically, `r n_transferred` players (`r pct_transferred`%) played for multiple clubs during the study period, justifying our cross-classified modeling approach.

[^6]: UEFA coefficient rankings are a points-based system used by European soccer's governing body to rank national associations and clubs based on their performance in European competitions, which determines qualification spots and seeding for tournaments like the Champions League and Europa League.

The primary outcome variable is the natural logarithm of the player's inflation-adjusted market value (in 2024 USD). Predictors are categorized into three distinct domains: (1) Player Characteristics, including age, height, primary position (Attacker, Midfielder, Defender, Goalkeeper), and citizenship status relative to the club ("Homegrown" vs. Foreign); (2) Performance and Status, captured by goals, assists, disciplinary records (red/yellow cards), and playing time; and (3) Club Context, specifically the proportion of foreign players on the roster and the club's UEFA coefficient, which serves as a proxy for recent competitive success and prestige. For further descriptive statistics on each variable see Appendix A, Table1.

## Cross-Classified Data Structure

Our data exhibit a cross-classified structure where players (*i*) are constant units, but club membership (*j*) changes over time (*t*). When a player transfers from club $j_1$ to club $j_2$, club-level predictors update to reflect the new club's characteristics. This structure requires modeling player and club as crossed random effects rather than nested effects. Figure 1 is a visual of a cross-classified model example of three players that either move through clubs or remain in the same club throughout their career.

```{r}
#| fig-height: 2.2
#| fig-width: 6.5

tibble(
  player = c(rep("Player A (Loyal)", 4),
             rep("Player B (Transfer)", 4),
             rep("Player C (Mobile)", 4)),
  year = rep(2020:2023, 3),
  club = c(
    "Real Madrid", "Real Madrid", "Real Madrid", "Real Madrid",
    "Dortmund", "Dortmund", "Man City", "Man City",
    "Ajax", "Juventus", "Juventus", "Bayern")
) |>
  ggplot(aes(x = year, y = player)) +
  geom_tile(aes(fill = club), color = "white", linewidth = 1, height = 0.65) +
  geom_text(aes(label = club), color = "white", fontface = "bold", size = 2.2) +
  scale_fill_viridis_d(option = "E", alpha=.8, begin = .8, end=.1) +
  scale_x_continuous(breaks = 2020:2023) +
  labs(title = "Figure 1. Cross-classified Data Structure Highlights How Players Move Between Clubs Over Time",
    x = "Season", y = NULL) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.text.y = element_text(face = "bold", size = 7)
  )

```

## Model Specification

We specify a cross-classified multilevel modelto partition the variance in market value across observations ($t$), players ($i$), and clubs ($j$). Crucially, because club attributes like internationalization and UEFA ranking vary year-to-year, they are modeled as Level 1 (time-varying) predictors rather than static Level 2 traits.

**Level 1: Observation Model (Time-Varying Effects)**

At the observation level, the log-transformed market value is modeled as a function of the player-club intercept ($\pi_{0ij}$), algebraic time trends, biological aging, and dynamic contextual metrics:

$$ \begin{aligned} \log(\text{Revenue})_{tij} &= \pi_{0ij} \\ &+ \pi_{1j}(\text{Year}_{t}) + \pi_{2j}(\text{Year}^2_{t}) + \pi_{3}(\text{Year}^3_{t}) \\ &+ \pi_{4}(\text{Age}_{it} - \bar{A}) + \pi_{5}(\text{Age}_{it} - \bar{A})^2 \\ &+ \pi_{6}(\text{Mins}_{Z,it}) \\ &+ \pi_{7}(\text{Foreign}_{tj}) + \pi_{8}(\text{UEFA}_{tj}) \\ &+ \mathbf{\Pi}_{perf} + \epsilon_{tij} \end{aligned} $$

*Note: The subscripts* ${tj}$ for Foreign and UEFA denote that these club attributes vary by time $t$.

**Level 2a: Player Model (Time-Invariant Traits)**

The player-club intercept ($\pi_{0ij}$) is decomposed into a fixed component based on static player characteristics and a random player effect ($u_{0i}$):

$$ \pi_{0ij} = \beta_{0j} + \gamma_{1}(\text{Height}_{i}) + \gamma_{2}(\text{Citizen}_{ij}) + \mathbf{\Gamma}_{demog} + u_{0i} $$

**Level 2b: Club Model (Contextual Trajectories)**

The club baseline ($\beta_{0j}$) and the time-varying slope parameters ($\pi_{1j}, \pi_{2j}$) include random effects to allow for heterogeneity in both growth velocity and trajectory curvature:

$$ \begin{aligned} \beta_{0j} &= \delta_{00} + v_{0j} \quad &\text{(Baseline Prestige)} \\ \pi_{1j} &= \delta_{10} + v_{1j} \quad &\text{(Linear Growth Deviation)} \\ \pi_{2j} &= \delta_{20} + v_{2j} \quad &\text{(Quadratic Curvature Deviation)} \\ \pi_{3} &= \delta_{30} \quad &\text{(Fixed Cubic Trend)} \end{aligned} $$

**Variance-Covariance Structure**

$$ u_{0i} \sim N(0, \sigma^2_{u}) \quad , \quad  \begin{pmatrix} v_{0j} \\ v_{1j} \\ v_{2j} \end{pmatrix} \sim N \left( \mathbf{0}, \begin{pmatrix} \sigma^2_{Int} & \sigma_{01} & \sigma_{02} \\ \sigma_{01} & \sigma^2_{Lin} & \sigma_{12} \\ \sigma_{02} & \sigma_{12} & \sigma^2_{Quad} \end{pmatrix} \right) $$

## Variables and Centering Strategies

To facilitate model convergence and ensure substantively meaningful intercepts, we applied specific centering and transformation strategies. Most notably, `Season Year` was centered at the 2020 pandemic baseline (2020=0), allowing us to interpret the intercept as the expected market value at the onset of the COVID-19 structural break. We introduced polynomial terms to capture theorized non-linear dynamics (see Figure 2): a cubic term for time to model the "Boom-and-Bust" market cycle (pre-pandemic inflation followed by post-pandemic deceleration), and a quadratic term for `Age` to empirically test the biological "Inverted-U" career trajectory. Additionally, `Minutes Played` was standardized (Z-scored) relative to each specific club-year to isolate a player's hierarchical status (e.g., Starter vs. Bench) independent of team-level rotation policies, while performance metrics were centered by position to account for role-specific expectations.

```{r}
#| fig-width: 8
#| fig-height: 4

y_title <- "Market Value, 2024 USD (log)"

# Define the specific years you want to show on the axis
selected_years <- c(2013, 2016, 2020, 2024) 

# Plot 1: Quadratic
p_1 <- clean_player_df |>
  ggplot(aes(x = season_year, y = usd_revenue_2024_log)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), 
              se = TRUE, color = "lightblue") +
  # Add a vertical line for the pandemic context
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50") +
  scale_x_continuous(breaks = selected_years) +
  labs(subtitle = "Quadratic Fit", # Subtitle here
       x = "", y = y_title) +
  theme_minimal()

# Plot 2: Cubic
p_2 <- clean_player_df |>
  ggplot(aes(x = season_year, y = usd_revenue_2024_log)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), 
              se = TRUE, color = "lightgreen") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50") +
  scale_x_continuous(breaks = selected_years) +
  labs(subtitle = "Cubic Fit", # Subtitle here
       x = "", y = "") + # Remove Y label for the second plot to save space
  theme_minimal()

# Combine with Patchwork
combined_plot <- p_1 + p_2 + 
  plot_annotation(
    title = "Figure 2: Market Value Impacted by Pandemic",
    caption = "Dashed line represents the 2020 Pandemic start."
  )

combined_plot
```

## Model Building and Statistical Software

We employed sequential model building: (1) null model for ICC estimation; (2) linear time polynomial; (3) cubic time polynomial; (4) random slope for quadratic time at club level[^7]; (5) player/club-season Level 1 predictors with quadratic age. Model selection was guided by likelihood ratio tests and information criteria. Model selection and comparison was guided by Likelihood Ratio Tests (LRT) to evaluate the statistical significance of nested model improvements. To balance goodness-of-fit against model complexity, we also consulted the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC). Models yielding significant LRT results (p\<.05) and lower values on both information criteria were retained for subsequent analysis. We describe each model in Appendix 2, Table 2 alongside ICC, AIC, and BIC values.

[^7]: We excluded player-level random slopes due to data sparsity; a majority of players appear in the dataset for only one or two seasons, making it mathematically impossible to disentangle their individual growth trajectories from residual error. In contrast, the club-level data provides a robust longitudinal structure with continuous observations across the decade, allowing for the stable estimation of club-specific market value trajectories.

All analyses were conducted in R version `r getRversion()` using the `lme4 1.1-37` package for mixed-effects model estimation with a Restricted Maximum Likelihood (REML) method. REML was selected to provide unbiased estimates of the variance components (random effects) by accounting for the degrees of freedom lost when estimating the fixed effects, ensuring that club and player heterogeneity were not underestimated.. To ensure robust model convergence given the complexity of the random effects structure, we utilized the BOBYQA (Bound Optimization BY Quadratic Approximation) optimizer. This algorithm is specifically designed to handle boundary constraints (such as variance components approaching zero) more effectively than the default Nelder-Mead optimizer. . Significance tests used Satterthwaite approximations via the `lmerTest 3.1-3` package.

# Results

The unconditional means (null) model revealed substantial clustering at the player level. Player-level variance accounted for 78% of the total variance ($ICC_{player}=.75$), while club-level variance accounted for 5% ($ICC_{club}=.03$),, highlighting that persistent individual heterogeneity dominates market valuations. Regarding temporal dynamics, model comparisons favored a cubic fixed effect for season year over a linear specification, $\chi^2(2)=45.66, p<.001$. We subsequently tested for club-specific heterogeneity in growth trajectories. A likelihood ratio test using a 50:50 mixture of $\chi^2_1$ ​ and $\chi^2_2$ ​ distributions[^8] confirmed that adding a random linear slope significantly improved model fit compared to the random-intercept model, $\chi^2(5)=97.13, p<.001$, indicating that clubs differ significantly not only in their growth rates but also in the curvature of their market value trajectories.[^9] Consequently, our final best-fitting model incorporates these complex random effects alongside time-varying player and club predictors ($\chi^2(19)=5626., p<.001$ compared to random slopes model). Model comparisons are shown in Appendix C, Table 1. Here we output our final model's variance components and fixed effects.

[^8]: We test whether a random slope with quadratic season year at the club level has better fit to the data. We find evidence supporting the need to include these random slopes using a likelihood ratio test with an alternative hypothesis, $H_A=\tau_{11} > 0$ since variance should be larger than 0. To test this hypothesis we use the likelihood ratio test (LRT) simply by $-2[\text{log likelihood}_{reduced} - \text{log likelihood}_{full}]$ to examine the difference. This code can be found in Appendix B.

[^9]: While we included random linear and quadratic slopes, a random cubic slope was excluded due to over-parameterization. Estimating the required 10 variance parameters with only 21 clubs led to singularity; therefore, the quadratic specification represents the maximal complexity supported by the data.

INCLUDE MODEL DIAGNOSTICS TEXT HERE (SHOW PLOTS IN APPENDIX

Table 1. Estimated Variance Components for Final Model

```{r}

# 1. Get Variance-Covariance Matrix (VCV) and convert to data frame
vc_summary <- as.data.frame(VarCorr(final_model))

# 2. Extract values based on Group and Component Name
# Note: var1 is the component (Intercept/Poly), var2 is NULL for variances
player_var <- vc_summary %>%
  filter(grp == "player_id") %>%
  pull(vcov)

club_int_var <- vc_summary %>%
  filter(grp == "club_id", var1 == "(Intercept)", is.na(var2)) %>%
  pull(vcov)

club_lin_var <- vc_summary %>%
  filter(grp == "club_id", var1 == "poly(season_year_c, 2, raw = TRUE)1", is.na(var2)) %>%
  pull(vcov)

club_quad_var <- vc_summary %>%
  filter(grp == "club_id", var1 == "poly(season_year_c, 2, raw = TRUE)2", is.na(var2)) %>%
  pull(vcov)

resid_var <- vc_summary %>%
  filter(grp == "Residual") %>%
  pull(vcov)

# 3. Create the final display table
vc_display <- tibble(
  Component = c(
    "Player (Reputation)", 
    "Club (Baseline Prestige)", 
    "Club (Linear Growth)", 
    "Club (Quadratic Curvature)", 
    "Residual (Unexplained)"
  ),
  Variance = c(player_var, club_int_var, club_lin_var, club_quad_var, resid_var),
  SD = sqrt(Variance)
)

# 4. Render
kable(vc_display, 
      digits = 4, 
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  column_spec(1, bold = TRUE)


```

Table 2. Fixed Effects Estimates from Final Model

```{r}

fx <- tidy(final_model, effects = "fixed", conf.int = TRUE) |>
  mutate(
    pct = (exp(estimate) - 1) * 100,
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    ),
    term_clean = case_when(
      term == "(Intercept)" ~ "Intercept",
      str_detect(term, "year_c.*1") ~ "Year (linear)",
      str_detect(term, "year_c.*2") ~ "Year²",
      str_detect(term, "year_c.*3") ~ "Year³",
      str_detect(term, "age_c.*1") ~ "Age (linear)",
      str_detect(term, "age_c.*2") ~ "Age²",
      term == "height_c" ~ "Height",
      term == "goals_c" ~ "Goals",
      term == "assists_c" ~ "Assists",
      term == "red_cards" ~ "Red cards",
      term == "yellow_cards" ~ "Yellow cards",
      term == "minutes_played" ~ "Minutes (Z)",
      term == "foreign_players_c" ~ "Foreign players %",
      term == "positionDefender" ~ "Defender",
      term == "positionGoalkeeper" ~ "Goalkeeper",
      term == "positionMidfield" ~ "Midfielder",
      term == "citizen_of_club_countryYes" ~ "Domestic citizen",
      term == "uefa_coefficient" ~ "UEFA Coefficient",
      term == "foreign_players" ~ "Proportion of Foregin Players",
      str_detect(term, "Region") ~ str_replace(term, "Region", ""),
      TRUE ~ term
    )
  )

fx_display <- fx |>
  filter(!str_detect(term_clean, "Intercept|Region|Asia|Africa|America|Middle")) |>
  select(term_clean, estimate, std.error, statistic, pct, sig)

kable(fx_display, digits = c(0, 3, 3, 2, 1, 0),
      booktabs = TRUE,
      col.names = c("Predictor", "β", "SE", "t", "% Change", "")) |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 10) |>
  footnote(general = "*** p < .001, ** p < .01, * p < .05",
           footnote_as_chunk = TRUE, threeparttable = TRUE)
```

### Market Evolution 

Using raw polynomials centered at 2020 ($t=0$), our model mathematically pinpoints the moment the market froze. The linear coefficient at the 2020 baseline is effectively zero ($\pi_1 = 0.00, p = 0.93$), indicating that at the exact onset of the pandemic, the relentless inflation of the previous decade came to a complete halt. The significant negative quadratic ($\pi_2 = -0.27$) and cubic ($\pi_3 = -0.12$) terms confirm the transition from a pre-2019 "Boom" into a structural correction. Figure 3 visualizes the complexity of club-level market dynamics by plotting the predicted trajectory for all clubs individually. This faceted view highlights how clubs diverge from the global trend. The vertical position of each chart confirms the initial hierarchy (e.g., random Intercepts), with elite clubs like Manchester City and Bayern Munich maintaining a consistently higher baseline than selling clubs (e.g., Sevilla, Porto). The distinct curvature of each line, particularly around the 2020 pandemic baseline, illustrates the significant variance in the random slopes. Clubs with concave trajectories (e.g., Juventus, Barcelona) plateaued or declined rapidly, confirming the risk of stagnation. Conversely, clubs with convex trajectories (e.g., Arsenal, Liverpool) show stronger momentum, indicating a more resilient recovery and active value generation throughout the decade.

```{r}
#| fig-width: 7
#| fig-height: 6
# 1. CREATE DATA FOR PLOTTING (Simple Centering, NOT Z-score)
# This ensures -7 = 2013 and +4 = 2024
plot_data_prep <- clean_player_df %>%
  mutate(season_year_c = season_year - 2020) %>%
  # Ensure other variables match what the formula expects
  mutate(
    minutes_played = as.numeric(scale(minutes_played)), # Z-score minutes
    age_c = Age - mean(Age, na.rm=TRUE) # Simple centering age
  )

# 2. REFIT MODEL ON THIS SPECIFIC DATA
# We must re-fit so the coefficients align with the -7 to +4 scale
Plot_Model <- lmer(
  usd_revenue_2024_log ~ 
    poly(season_year_c, 3, raw = TRUE) + # Use raw for plotting model
    height_c + position + citizen_of_club_country + Region +
    poly(age_c, 2, raw = TRUE) + 
    goals + assists + red_cards + yellow_cards + minutes_played +
    # Note: Model 3.2 structure (No Foreign/UEFA fixed effects)
    (1 | player_id) + (poly(season_year_c, 2, raw = TRUE) | club_id),
  data = plot_data_prep, 
  control = lmerControl(optimizer = "bobyqa")
)

# 3. Create Metadata Dictionary
club_meta <- plot_data_prep %>%
  mutate(club_id = as.character(club_id)) %>%
  group_by(club_id) %>%
  summarise(Club = first(Club)) %>% 
  ungroup()

# 4. Generate Predictions (Full Range)
# using [all] on the REFITTED model will now span -7 to +4
club_pred <- ggpredict(Plot_Model, 
                       terms = c("season_year_c [all]", "club_id"), 
                       type = "random") 

# 5. Process Data
final_plot_data <- club_pred %>%
  as.data.frame() %>%
  mutate(group = as.character(group)) %>% 
  left_join(club_meta, by = c("group" = "club_id")) %>%
  group_by(Club) %>%
  mutate(max_val = max(predicted)) %>%
  ungroup() %>%
  mutate(Club = reorder(Club, -max_val))

# 6. Plot
ggplot(final_plot_data, aes(x = x + 2020, y = predicted)) +
  
  # Pandemic Line
  geom_vline(xintercept = 2020, linetype = "dashed", 
             color = "red", alpha=0.5) +
  
  # Trajectories
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "#440154", alpha = 0.1) +
  geom_line(color = "#440154", linewidth = 1.2) +
  
  # Facets
  facet_wrap(~Club, ncol = 5) + 
  
  # Ensure axis covers full range
  scale_x_continuous(breaks = seq(2013, 2024, 2), 
                     limits = c(2013, 2024)) +
  
  labs(
    title = "Figure 3. Club Trajectories Through the Pandemic (2013-2024)",
    subtitle = "Predicted Log Market Value, Dash Line Begining of Pandemic",
    x = "Season Year",
    y = "Log Market Value (2024 USD)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Biological Age

Our model predicts the peak age to be 26.9[^10] for optimal transfer market value we visualize in Figure 4 the biological constraints on valuation. The curve rises steeply as players transition from 'prospects' (Age 17–23), flattens into a 'prime' plateau between ages 26 and 28, and declines steadily thereafter. The model predicts that a player's value drops by approximately 1.5 log points (from a peak of 17.0 to 15.5) as they age from 27 to 35, reflecting the market's heavy discount on older assets with limited resale potential.

[^10]: We use the delta method, specifically Taylor series approximation, to propagate uncertainty through nonlinear transformation of $-\beta_{age} / (2 \times \beta_{age^2})$.

```{r}
# average age
mean_age <- clean_player_df |> reframe(mean(Age, na.rm=TRUE)) |> pull()

# Simulate from the estimated sampling distribution
# "-b1/(2*b2)" Report with confidence interval using delta method

sims <- sim(final_model, n.sims = 1000) 

peak_sims <- sims@fixef |> 
  as.data.frame() |> 
  tibble() |> 
  reframe( mean(
    (`poly(age_c, 2, raw = TRUE)1` * -1) / 
      (2 * `poly(age_c, 2, raw = TRUE)2`) )
    ) |> 
  pull()



# Empirical CI
#cat("Peak Age = ", round(mean_age + peak_sims, 2))

# The delta method uses calculus (specifically, Taylor series approximation) 
## to propagate uncertainty through nonlinear transformations.
```

```{r}
# 1. Get Mean Age to reverse the centering
# We need the exact mean used in the dataset
mean_age <- mean(clean_player_df$Age, na.rm = TRUE)

# 2. Generate Predictions
pred_age <- ggpredict(final_model, terms = "age_c [all]") 

# 3. Back-Transform X-axis
plot_data_age <- pred_age %>%
  as.data.frame() %>%
  mutate(
    # Formula: Real_Age = Centered_Age + Mean_Age
    age_real = x + mean_age
  )

# 4. Plot with Real Ages
ggplot(plot_data_age, aes(x = age_real, y = predicted)) +
  
  # Trend Line and Ribbon
  geom_line(color = "#3b528b", linewidth = 1.5) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "#3b528b", alpha = 0.2) +
  
  # Add Peak Reference Line (Visual aid)
  geom_vline(xintercept = 27, linetype = "dashed", color = "gray50") +
  annotate("text", x = 27.5, y = min(plot_data_age$predicted) + 1, 
           label = "Peak ~27", hjust = 0, color = "gray40", size = 3.5) +
  
  # formatting
  scale_x_continuous(breaks = seq(16, 40, 4)) + # Show ages 16, 20, 24...
  
  labs(
    title = "Figure 4. The Career Arc, Predicted Log Market Value by Age",
    x = "Player Age (Years)", 
    y = "Log Market Value"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

### **Contextual Premiums and Discounts**

Squad status remains the dominant behavioral driver. The standardized coefficient for minutes played ($\pi_6 = 0.35$) indicates a 42% valuation premium ($e^{0.35}-1$) for key starters (+1 SD) compared to average players. Conversely, we observe a robust "Domestic Discount," where homegrown citizens are valued approximately 17% lower ($\gamma_2 = -0.19$) than comparable foreign imports. While we hypothesized a "Diversity Premium" for clubs with higher international representation, the fixed effect for `foreign_players` was positive but not statistically significant ($\pi_7 = 0.10, p = 0.32$) in the final model. Similarly, the `uefa_coefficient` showed no significant effect ($p = 0.70$). This suggests that once we account for the massive heterogeneity in Club Random Effects (i.e., Baseline Prestige and Growth Trajectories) and individual player reputation ($ICC \approx 0.78$), specific club-level metrics add little unique explanatory power. In Figure 5, we present relative market valuation for Level 2 fixed effects (e.g., citizenship, region origin, and position). The coefficient for "Homegrown (Citizen)" falls strictly to the left of zero, confirming that domestic players are valued significantly lower than comparable foreign imports ($\beta=−0.19$), potentially reflecting the recruitment premium required to attract international talent. Attackers (the baseline) command the highest market value. All other positions show significant negative effects, with Goalkeepers facing the steepest penalty ($\beta=−1.06$), suggesting the market disproportionately rewards offensive output. While most regional coefficients overlap with zero (i.e., indicating no difference from the European baseline), Latin America stands out as the only region with a positive premium ($\beta=0.19$). This suggests that South American players function as a "premium export brand" in the global market, distinct from other non-European origins.

```{r}
plot_summs(final_model, 
           coefs = c("Homegrown (Citizen)" = "citizen_of_club_countryYes",
                     "Defender" = "positionDefender",
                     "Midfielder" = "positionMidfield",
                     "Goalkeeper" = "positionGoalkeeper",
                     "Latin America" = "RegionLatin America & Caribbean",
                     "Asia & Pacific" = "RegionEast Asia & Pacific",
                     "North America" = "RegionNorth America",
                     "Middle East & N. Africa" = "RegionMiddle East & North Africa",
                     "Sub-Saharan Africa" = "RegionSub-Saharan Africa"),
           colors = "#440154", # Viridis Purple
           point.shape = 18) + 
  
  labs(title = "Figure 5. Valuation Gaps Across Citizenship, Position, and Region.",
       subtitle = "Discounts (Left) vs. Premiums (Right) relative to baseline",
       x = "Effect on Log Market Value") +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

### EXPAND THE DISCUSSION SECTION A BIT MORE

# Discussion

This study demonstrates that player market value in elite football is driven primarily by persistent individual reputation but is meaningfully moderated by club context and market cycles. Our cross-classified approach properly attributes variance to portable player effects versus transient club effects.

**Market Cycles.** The cubic temporal trend reveals distinct boom-bust phases rather than constant market growth. The rapid inflation period (2015--2019) following the Neymar transfer record and subsequent post-pandemic correction highlight the cyclical nature of transfer market valuations.

**Biological Constraints.** The strict inverted-U age relationship, with peak value around age 27, confirms that physical peak represents the optimal balance between experience and athleticism, with practical implications for optimal transfer timing.

**Squad Status Dominance.** Minutes played emerged as the strongest predictor, with starters valued over 40% higher than bench players. This universal signal transcends position-specific metrics, suggesting that coaches' implicit assessments are incorporated into market valuations.

**Contextual Premiums.** The diversity premium for internationally-oriented clubs and domestic discount for homegrown players suggest systematic biases that could be exploited for competitive advantage.

## Limitations

Several limitations warrant acknowledgment. First, our sample focuses on elite clubs, limiting generalizability. Second, minutes played is likely endogenous; our estimates should be interpreted as associations rather than causal effects. Third, we constrained higher-order time terms to be fixed due to convergence issues. Fourth, residual diagnostics revealed minor departures from normality in tails, though our large sample ensures robust estimation.

### Conclusion

By properly modeling the cross-classified structure where players transfer between clubs, we quantify how individual performance, biological constraints, and club context jointly shape valuations. The substantial player-level variance confirms that persistent individual reputation dominates, while club context creates meaningful premiums and trajectories. Future research should extend this framework to lower-tier leagues and employ instrumental variable approaches to address endogeneity.

\newpage

# References {.unnumbered}

Deloitte. (2021). *Riding the challenge: Annual Review of Football Finance 2021*. Deloitte Sports Business Group. <https://www2.deloitte.com/uk/en/pages/sports-business-group/articles/annual-review-of-football-finance-europe.html>

Drewes, M., Daumann, F., & Follert, F. (2021). Exploring the sports economic impact of COVID-19 on professional soccer. *Soccer & Society, 22*(1-2), 125-137. <https://doi.org/10.1080/14660970.2020.1802256>

Herm, S., Callsen-Bracker, H. M., & Kreis, H. (2014). When the crowd evaluates soccer players' market values: Accuracy and evaluation attributes of an online community. *Sport Management Review, 17*(4), 484-492. <https://doi.org/10.1016/j.smr.2013.12.006>

Poli, R., Ravenel, L., & Besson, R. (2017). *Transfer market analysis* (CIES Football Observatory Monthly Report No. 27). CIES Football Observatory. <https://football-observatory.com/IMG/sites/mr/mr27/en/>

\clearpage

\newpage

# Appendix A

## Table 1: Variable Descriptions and Descriptive Statistics (*N*=7,168 raw) {.unnumbered}

```{r}

# 1. Define Variable Metadata (Continuous Only)
var_metadata <- tribble(
  ~Variable, ~"Model Variable", ~Description, ~"Transformation Strategy",
  "usd_revenue_2024", "log_revenue", "Market Value (2024 USD)", "Log-transformed (ln)",
  "season_year", "season_year_c", "Season Year", "Centered at 2020 (0 = 2020)",
  "Age", "age_c", "Player Age (Years)", "Grand-mean centered + Quadratic",
  "height_in_cm", "height_c", "Height (cm)", "Grand-mean centered",
  "minutes_played", "minutes_z", "Minutes Played", "Z-scored within Club-Year",
  "goals", "goals_c", "Goals Scored", "Centered by Position Mean",
  "assists", "assists_c", "Assists Made", "Centered by Position Mean",
  "foreign_players", "foreign_c", "Club Foreign %", "Grand-mean centered",
  "uefa_coefficient", "uefa_c", "UEFA Coefficient", "Year-mean centered",
  "red_cards", "red_cards", "Red Cards", "Raw count",
  "yellow_cards", "yellow_cards", "Yellow Cards", "Raw count"
)

# 2. Calculate Continuous Stats
stats_cont <- dat %>%
  summarise(
    across(c(usd_revenue_2024, season_year, Age, height_in_cm, 
             minutes_played, goals, assists, foreign_players, uefa_coefficient,
             red_cards, yellow_cards),
           list(Mean = ~mean(., na.rm = TRUE), SD = ~sd(., na.rm = TRUE)),
           .names = "{.col}___{.fn}")
  ) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), 
               names_sep = "___") %>%
  pivot_wider(names_from = Stat, values_from = value) %>%
  mutate(
    Stat_String = case_when(
      Variable == "usd_revenue_2024" ~ sprintf("$%.1fM (%.1fM)", Mean/1e6, SD/1e6),
      Variable == "foreign_players" ~ sprintf("%.1f%% (%.1f%%)", Mean*100, SD*100),
      TRUE ~ sprintf("%.2f (%.2f)", Mean, SD)
    )
  ) %>%
  select(Variable, Stat_String)

# 3. Calculate Detailed Categorical Stats (Helper Function)
get_cat_stats <- function(df, col_name, pretty_name) {
  df %>%
    count(.data[[col_name]]) %>%
    mutate(
      Prop = n / sum(n) * 100,
      Stat_String = sprintf("%s (%.1f%%)", n, Prop),
      Variable = as.character(.data[[col_name]]), 
      "Model Variable" = "",
      Description = paste(" -", Variable), 
      "Transformation Strategy" = ""
    ) %>%
    select(Description, Stat_String, "Transformation Strategy")
}

# Generate rows for each categorical variable
cat_rows <- bind_rows(
  tibble(Description = "Citizenship (Homegrown)", Stat_String = "", `Transformation Strategy` = "Binary (1=Citizen)"),
  get_cat_stats(dat, "citizen_of_club_country", "Citizenship"),
  
  tibble(Description = "Primary Position", Stat_String = "", `Transformation Strategy` = "Factor (Ref: Attack)"),
  get_cat_stats(dat, "position", "Position"),
  
  tibble(Description = "Region of Origin", Stat_String = "", `Transformation Strategy` = "Factor (Ref: Europe)"),
  get_cat_stats(dat, "Region", "Region")
)

# 4. Merge and Render
# First merge continuous metadata with stats
table_cont <- var_metadata %>%
  left_join(stats_cont, by = "Variable") %>%
  rename(`Raw Stats (Mean/SD)` = Stat_String) %>%
  select(Description, `Raw Stats (Mean/SD)`, `Transformation Strategy`)

# Rename categorical columns to match for binding
cat_rows <- cat_rows %>% rename(`Raw Stats (Mean/SD)` = Stat_String)

# Bind everything
final_table <- bind_rows(table_cont, cat_rows)

# Render with caption = NULL to remove "Table 3"
kable(final_table, 
      format = "latex", 
      booktabs = TRUE, 
      caption = NULL) %>%
kable_styling(latex_options = c("scale_down", "hold_position"), 
              font_size = 10) %>%  column_spec(1, bold = TRUE, width = "5cm") %>%
  column_spec(2, width = "4cm") %>%
  column_spec(3, italic = TRUE, width = "5cm") %>%
  pack_rows("Continuous Variables", 1, 11) %>%
  pack_rows("Categorical Variables", 12, nrow(final_table))
```

\newpage

# Appendix B

## Annotated Model Syntax

```{r}
#| eval: false
#| echo: true

#______________________________________________________________________________
# MODEL SPECIFICATION & FITTING
#______________________________________________________________________________

# 1. MODEL NULL (Unconditional Means)
# Purpose: Partition variance (calculate ICC) between Player and Club levels.
model_null <- lmer(usd_revenue_2024_log ~ (1|player_id) + (1|club_id), 
                   data = clean_player_df, REML = TRUE, control = strict_control)

# 2. MODEL LINEAR TIME TREND
# Purpose: Test for general linear inflation in market values 
## (Base model for time).
model_time_linear <- lmer(usd_revenue_2024_log ~ season_year_c + 
                          (1|player_id) + (1|club_id),
                          data = clean_player_df, REML = TRUE, control = strict_control)

# 3. MODEL CUBIC TIME TREND (Fixed Effects)
# Purpose: Test for non-linear "Boom and Bust" cycles. 
## (raw=TRUE for interpretability)
model_time_cubic <- lmer(usd_revenue_2024_log ~ poly(season_year_c, 3, raw=TRUE) + 
                         (1|player_id) + (1|club_id),
                         data = clean_player_df, REML = TRUE, control = strict_control)

# 4. MODEL RANDOM QUADRATIC SLOPE (Club Level)
# Purpose: Test if trajectory curvature varies by club. 
## (Linear slope model omitted for brevity)
model_slope_quadratic <- lmer(usd_revenue_2024_log ~ poly(season_year_c, 3, raw=TRUE) + 
                              (1|player_id) + (poly(season_year_c, 2, raw=TRUE) | club_id),
                              data = clean_player_df, REML = TRUE, control = strict_control)

# 5. MODEL (FINAL): FULL SPECIFICATION
# Purpose: Estimate marginal effects of performance, status, 
## and context controlling for time.
model_player <- lmer(
  usd_revenue_2024_log ~ 
    poly(season_year_c, 3, raw=TRUE) +          # Temporal Controls
    poly(age_c, 2, raw=TRUE) +                  # Career Arc
    height_c + position + Region +              # Player Traits
    citizen_of_club_country +                   # Context
    goals + assists + red_cards + yellow_cards + minutes_played + # Performance
    uefa_coefficient + foreign_players +        # club fixed effects
    (1|player_id) + (poly(season_year_c, 2, raw=TRUE) | club_id), # Random Effects
  data = clean_player_df, REML = TRUE, control = strict_control
)

#______________________________________________________________________________
# LIKELIHOOD RATIO TEST (MIXTURE DISTRIBUTION)
#______________________________________________________________________________
# Corrects p-values for testing variance components on the boundary (Stram & Lee, 1994)
# LRT test for 2 random slopes
calculate_lrt_mixture <- function(reduced_model, full_model) {
  
  # 1. Extract Log-Likelihoods
  ll_reduced = as.numeric(logLik(reduced_model))
  ll_full = as.numeric(logLik(full_model))
  
  # 2. Calculate LRT Statistic
  lrt_stat = -2 * (ll_reduced - ll_full)
  
  # 3. Calculate Difference in Parameters (df)
  df_reduced = attr(logLik(reduced_model), "df")
  df_full = attr(logLik(full_model), "df")
  df_diff = df_full - df_reduced
  
  # 4. Calculate P-Value based on the difference
  
  if (df_diff == 3) {
    # --- NEW BLOCK FOR QUADRATIC SLOPE ---
    # Going from 2 Random Effects (Int+Lin) -> 3 Random Effects (Int+Lin+Quad)
    # We add 1 variance + 2 covariances = 3 parameters.
    # The mixture is 0.5 * ChiSq(2) + 0.5 * ChiSq(3)
    p_val = 0.5 * pchisq(lrt_stat, df = 2, lower.tail = FALSE) + 
      0.5 * pchisq(lrt_stat, df = 3, lower.tail = FALSE)
    
    msg = "Mixture of ChiSq(2) and ChiSq(3)"
    
  } else if (df_diff == 2) {
    # Going from 1 Random Effect (Int) -> 2 Random Effects (Int+Lin)
    # Adds 1 variance + 1 covariance = 2 parameters
    p_val = 0.5 * pchisq(lrt_stat, df = 1, lower.tail = FALSE) + 
      0.5 * pchisq(lrt_stat, df = 2, lower.tail = FALSE)
    
    msg = "Mixture of ChiSq(1) and ChiSq(2)"
    
  } else if (df_diff == 1) {
    # Adding a single parameter (e.g. just a variance, no covariance)
    p_val = 0.5 * pchisq(lrt_stat, df = 0, lower.tail = FALSE) + 
      0.5 * pchisq(lrt_stat, df = 1, lower.tail = FALSE)
    
    msg = "Mixture of ChiSq(0) and ChiSq(1)"
    
  } else {
    # Fallback for standard fixed effects comparisons
    p_val = pchisq(lrt_stat, df = df_diff, lower.tail = FALSE)
    msg = paste0("Standard ChiSq(", df_diff, ")")
  }
  
  # Output
  cat("LRT Statistic:", round(lrt_stat, 4), "\n")
  cat("Parameter Diff:", df_diff, "\n")
  cat("Distribution:  ", msg, "\n")
  cat("P-Value:       ", format.pval(p_val, eps = .0001), "\n")
}

```

```{r}
calculate_lrt_mixture(model_time_cubic, model_slope_quadratic)
```

\newpage

# Appendix C

## Table 1. Model Fit Comparison (AIC, BIC, and ICC) {.unnumbered}

|  |  |  |  |  |
|-------------|--------------------------------|----------|---------|----------|
| **Models** | **Purpose & Research Justification** | ICCplayer/club | AIC | BIC |
|  **Null** | Calculate ICC and partition variance between players and clubs before adding predictors | .75/.03 | 20105.19 | 20132.62 |
| **Time Linear** | Test for a general linear trend in market value growth | .76/.03 | 20050.17 | 20084.46 |
| **Time Cubic** | Test for non-linear market dynamics (e.g., acceleration or deceleration/plateaus) | .76/.03 | 20021.61 | 20069.61 |
| **Random Slope Quadratic** | Test if the growth curvature varies by club | .76/.04 | 19934.48 | 20016.76 |
| **Varying Level 1 fixed effects** | Estimate the marginal value of performance metrics and player/club characteristics controlling for time | .78/.05 | 14467.31 | 14679.87 |
| **Final Model Selection** | Compare models using LRT, AIC/BIC, and residual diagnostics (QQ plots) to select the most parsimonious and robust fit |  |  |  |

### PRESENT MODEL DIAGNOSTICS BELOW

```{r}

# slopes
qqnorm(ranef(final_model)$player_id[,1]) # group factor
qqline(ranef(final_model)$player_id[,1])


qqnorm(ranef(final_model)$club_id[,1]) # group factor
qqline(ranef(final_model)$club_id[,1])

qqnorm(ranef(final_model)$club_id[,2]) # group factor
qqline(ranef(final_model)$club_id[,2])

re <- ranef(final_model)$club_id

# 2. Rename columns for clarity (poly names are messy)
# Note: Ensure your model actually has 3 REs. If not, adjust indices.
colnames(re) <- c("Intercept", "Linear_Slope", "Quad_Slope")

# 3. Create a Scatterplot Matrix (Pairs Plot)
# This plots every random effect against every other random effect
pairs(re, 
      main = "Correlation between Random Effects",
      pch = 19, 
      col = "dodgerblue",
      lower.panel = panel.smooth, # Adds trend lines automatically
      upper.panel = NULL)         # Hides duplicate upper half


```

```{r}
#| fig-width: 7
#| fig-height: 10
check_model(final_model)
```

```{r}
rm(model_null)
rm(model_time_linear)
rm(model_time_cubic)
rm(model_slope_quadratic)
rm(model_player)
rm(final_model)
rm(dat) 
rm(clean_player_df)
gc()
```
