---
title: "FinalProject_SURV687"
author: "Jianing Zou"
format: pdf
editor: visual
---

```{r}
library(arrow)
library(lme4)
library(lmerTest)    
library(dplyr)
library(performance) 
library(broom.mixed) 
library(jtools) # new package

options(scipen = 999)
```

```{r}
# Read .parquet file
data <- read_parquet("player_df.parquet")

# View the data
glimpse(data)

```

# Data Structure

Level 1 (time): season_year: season-year within player Level 2 (player): name: players Level 3 (club): club_id: clubs

L1 time-varying: season_year, yellow_cards, red_cards, goals, assists, minutes_played L2 player-invariant: country_of_birth, sub_position, position, height_in_cm L3 club-invariant: season_year

# Multilevel Model (Nested)

-   Look at this [link ](https://rpsychologist.com/r-guide-longitudinal-lme-lmer)

Do we need centering for time-varying covariates so L1 effects are “purely within-person"?

Where to put seanson_year?

Do we finally make the decision that we use nested model not cross-classfied model?

We need to also consider whether we need to put random slopes and interaction term or not. If we do, then which variable need to be considered?

1.  only random intercept model.baseline
2.  add new variables step by step.

```{r}

player_df <- data %>%
  mutate(
    player_id = factor(player_id),
    club_id   = factor(club_id)
  )
```

-   Intercept only model: We want to model `r usd_revenue_2024` because market value is accounting for inflation and converted to 2024 US dollars.

```{r}
intercept_only <-  lmer(
  usd_revenue_2024 ~ (1 | club_id) +(1|club_id:player_id),  
  data = player_df, REML = TRUE)

summ(intercept_only)
```

-   Noticed that the club id only accounts for 7% of the variance in market value, while the player id accounts for 53% as expected. But what happens to the ICC if we model these data on the log scale?

```{r}
intercept_only_log <-  lmer(
  log(usd_revenue_2024) ~ (1 | club_id) +(1|club_id:player_id),  
  data = player_df, REML = TRUE)

summ(intercept_only_log)
```

-   As expected, logged US dollars tends to normalize values, and our ICC jumped to 81%, while the AIC and BIC dropped.

```{r}
compare_performance(intercept_only, intercept_only_log,
                    metrics = c("AIC", "BIC", "R2", "ICC", "RMSE"))
```

### Including predictors

```{r}

mA <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year + yellow_cards + red_cards + goals + assists +
    # Level-2 (player-invariant)
    Region +  position + height_in_cm + citizen_of_club_country +
    # Level-3 (club-invariant)
    foreign_players +
    # Random intercepts: club and player-within-club
   (1 | club_id) + (1|club_id:player_id),
  data = player_df, REML = TRUE
)

summ(mA)

compare_performance(intercept_only_log, mA,
                    metrics = c("AIC", "BIC", "R2", "ICC", "RMSE"))
```

-   NEXT, figure out which variables to center overall and by club such as (height - mean(height) by club and or year. The idea is to help us interpret the coefficients.

-   Think about other variables to include. Let me know which ones and I can add them to the dataframe such as average height by club X year.
