---
title: "FinalProject_SURV687"
author: "Jianing Zou"
format: pdf
editor: visual
---

```{r}
library(arrow)
library(lme4)
library(lmerTest)    
library(dplyr)
library(performance) 
library(broom.mixed) 

```
```{r}

# Read .parquet file
data <- read_parquet("player_df.parquet")

# View the data
head(data)

```




# Data Stricture

Level 1 (time): season_year: season-year within player
Level 2 (player): name: players
Level 3 (club): club_id: clubs

L1 time-varying: season_year, yellow_cards, red_cards, goals, assists, minutes_played
L2 player-invariant: country_of_birth, sub_position, position, height_in_cm
L3 club-invariant: season_year


# Multilevel Model (Nested)

Do we need centering for time-varying covariates so L1 effects are “purely within-person"?

Where to put seanson_year?

Do we finally make the decision that we use nested model not cross-classfied model?

We need to also consider whether we need to put random slopes and interaction term or not. If we do, then which variable need to be considered?



1. only random intercept model.baseline
2. add new variables step by step.

```{r}

player_df <- data %>%
  mutate(
    player_id = factor(player_id),
    club_id   = factor(club_id)
  )
```

```{r}

mA <- lmer(
  market_value_ave_usd ~ 
    # Level-1 (within-player, time-varying)
    season_year + yellow_cards + red_cards + goals + assists +
    # Level-2 (player-invariant)
    player_id +  country_of_birth + sub_position + position + height_in_cm +
    # Level-3 (club-invariant)
    club_id + 
    # Random intercepts: club and player-within-club
   (1 | club_id) + (1 | player_id),
  data = player_df, REML = TRUE
)
summary(mA)

```

