---
title: "FinalProject_SURV687"
author: "Jianing Zou"
format: pdf
editor: visual
---

```{r}
library(arrow)
library(lme4)
library(lmerTest)    
library(dplyr)
library(performance) 
library(broom.mixed) 
library(jtools) # new package

options(scipen = 999)
```

```{r}
# Read .parquet file
data <- read_parquet("player_df.parquet")

# View the data
glimpse(data)

```

# Data Structure

Level 1 (time): season_year: season-year within player Level 2 (player): name: players Level 3 (club): club_id: clubs

L1 time-varying: season_year, yellow_cards, red_cards, goals, assists, minutes_played L2 player-invariant: country_of_birth, sub_position, position, height_in_cm L3 club-invariant: season_year

seanson_year: 2013, 2014, 2015, 2016, 2017, 2018, 2020, 2021, 2022, 2023, 2024.

position: Attack, Defender, Goalkeeper, Midfield.

Region: East Asia & Pacific, Europe & Central Asia, Latin America & Caribbean, Middle East & North Africa, North America, Sub-Saharan Africa.

citizen_of_club_country: YES, NO.


```{r}

player_df <- data %>%
  mutate(
    player_id = factor(player_id),
    club_id   = factor(club_id),
    position = factor(position),
    Region = factor(Region),
    citizen_of_club_country = factor(citizen_of_club_country)
    
  )
```

-   Check missing values.
```{r}
# Which variable has missing value
colSums(is.na(player_df))

# Which observation has missing values
player_df %>%
  filter(!complete.cases(.))

# Exclude all missing values, only 7023 observations now
clean_player_df <- na.omit(player_df)

```

-   Intercept only model: We want to model `r usd_revenue_2024` because market value is accounting for inflation and converted to 2024 US dollars.


# Fit the model

## Cross or nested?

-   The player went to different clubs across years, thus they are cross-clubs, not nested in one club.

-   So we use (1 | club_id) + (1|player_id) here, not (1 | club_id) + (1|club_id : player_id) here, since the latter one is used only when player are nested in the clubs.

```{r}


clean_player_df %>%
  distinct(player_id, club_id) %>%      
  count(player_id, name = "num_clubs") %>%  
  filter(num_clubs > 1) %>%              
  arrange(desc(num_clubs))

```


```{r}
intercept_only <-  lmer(
  usd_revenue_2024 ~ (1 | club_id) +(1|player_id),  
  data = clean_player_df, REML = TRUE)

summ(intercept_only)
```


# Multilevel Model (Nested)

-   Look at this [link](https://rpsychologist.com/r-guide-longitudinal-lme-lmer)

Do we need centering for time-varying covariates so L1 effects are “purely within-person"?

Where to put seanson_year?

Do we finally make the decision that we use nested model not cross-classfied model?

We need to also consider whether we need to put random slopes and interaction term or not. If we do, then which variable need to be considered?

1.  only random intercept model.baseline
2.  add new variables step by step.


-   Noticed that the club id only accounts for 7% of the variance in market value, while the player id accounts for 53% as expected. But what happens to the ICC if we model these data on the log scale?

```{r}
intercept_only_log <-  lmer(
  log(usd_revenue_2024) ~ (1 | club_id) + (1|player_id),  
  data = clean_player_df, REML = TRUE)

summ(intercept_only_log)
```

-   As expected, logged US dollars tends to normalize values, and our ICC jumped to 81%, while the AIC and BIC dropped.

```{r}
compare_performance(intercept_only, intercept_only_log,
                    metrics = c("AIC", "BIC", "R2", "ICC", "RMSE"))
```

### Centering


-   To make our model interpretable, we need to center age and height variable

-   Center age and height_in_cm variable here.

```{r}

clean_player_df <- clean_player_df %>%
  mutate(
    mean_age = mean(Age),
    mean_height = mean(height_in_cm),
    age_cen  = Age - mean_age,
    height_cen = height_in_cm - mean_height
  )
```

### Including predictors


#### Add position and season year

-   Both AIC and BIC of model adding position and season year are smaller than that of intercept-only model.

```{r}
# Orginal model. Set REML = FALSE to compare fixed effects
intercept_only_log <-  lmer(
  log(usd_revenue_2024) ~ (1 | club_id) + (1|player_id),  
  data = clean_player_df, REML = FALSE)

m1 <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +
    # Level-2 (player-invariant)
    position +
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)

summ(m1)

anova(intercept_only_log, m1)

```

#### Add centering age and centering height

-   After adding age, AIC and BIC 

```{r}

m2 <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year + 
    # Level-2 (player-invariant)
    position + age_cen + height_cen +
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)

summ(m2)

anova(m1,m2)

```
#### Add yellow_cards, red_cards

-   The model adding only yellow cards fit the data better, because both AIC and BIC are smallest among all three models here.

```{r}

m3 <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + 
    # Level-2 (player-invariant)
    position + age_cen + height_cen +
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)

m4 <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + red_cards +
    # Level-2 (player-invariant)
    position + age_cen + height_cen +
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)



anova(m2,m3,m4)
```

#### Add goals, assists

-   The model adding goals and assists fit the data better

```{r}
m5 <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + goals + 
    # Level-2 (player-invariant)
    position + age_cen + height_cen +
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)

m6 <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + goals + assists +
    # Level-2 (player-invariant)
    position + age_cen + height_cen +
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)


anova(m3,m5,m6)

summ(m6)
  
```

#### Adding Region and citizen_of_club_country

-   Adding region and citizen_of_club_country yield better AIC and BIC.

```{r}

m7 <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + goals + assists +
    # Level-2 (player-invariant)
    position + age_cen + height_cen + Region + 
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)

m8  <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + goals + assists +
    # Level-2 (player-invariant)
    position + age_cen + height_cen + Region + citizen_of_club_country + 
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)

anova(m6,m7,m8)

```

#### Adding foreign_players

-   m8 has higher AIC but lower BIC.

-   m9 has lower AIC but lower BIC.

-   By considering both good fit and complexity, we choose m8, the model without foreign_players.

```{r}

m9  <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + goals + assists +
    # Level-2 (player-invariant)
    position + age_cen + height_cen + Region + citizen_of_club_country + 
    foreign_players +
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = FALSE
)

anova(m8,m9)
```


#### Final model

```{r}



m_final  <- lmer(
  log(usd_revenue_2024)  ~ 
    # Level-1 (within-player, time-varying)
    season_year +  yellow_cards + goals + assists +
    # Level-2 (player-invariant)
    position + age_cen + height_cen + Region + citizen_of_club_country + 
    # Random intercepts: club and player-within-club
   (1 | club_id) + (1| player_id),
  data = clean_player_df, REML = TRUE
)


summ(m_final)
```

-   NEXT, figure out which variables to center overall and by club such as (height - mean(height) by club and or year. The idea is to help us interpret the coefficients.

-   Think about other variables to include. Let me know which ones and I can add them to the dataframe such as average height by club X year.


