---
title: "Model Fit"
author: "Money Ballers"
date: "today"
embed-resources: true
editor: visual
execute:
  warning: false
  error: false
  message: false
format: pdf
---

```{r}
pacman::p_load(arrow, lme4, lmerTest, performance, broom, broom.mixed, patchwork,
               arm, viridis, knitr, jtools, ggeffects, ggdist, 
               gt, tidyverse)

options(scipen = 999)

# Set theme for plots
theme_set(theme_minimal())
```

```{r}

# Setting repository path
#repo_path <- "/Users/namomac/Desktop/Moneyball_FC/inputs/"
repo_path <- "D:/repos/Moneyball_FC/inputs/"

# Reading data
dat <- read_parquet(str_c(repo_path, "player_df.parquet")) 
#dat <- read_parquet("player_df.parquet") 

```

### **Time** **Centering**

-   **Year**: Centering at the start of the COVID-19 pandemic anchors the model to the most significant economic shock of the decade. This makes the intercept ($\beta_0$) interpretable as the expected market value during the 2020 season, rather than an arbitrary start date (2013). Mathematically, placing "0" near the center of the time series reduces collinearity between the linear, quadratic, and cubic time terms, improving model stability.

    -   Q: How does market value change annually relative to the 2020 pandemic baseline?

### **Grand Mean Centering**

-   **Age**: Market value is determined by biological career stage (e.g., "is Messi 29 and in his prime?" vs "is Messi 19 and a prospect?"). It does *not* matter if you are the oldest guy on a young team (relative age); the market still sees you as 29. We include the linear term (`age_c`) to catch the general trend and the squared term (`age_sq`) to capture the peak and decline curve.

    -   Q: How does market value change as a player moves through their biological career arc (from prospect to prime to veteran)?

-   **Foreign Player Proportion**: Since no club has 0% foreign players (lowest is 28%), we shift the zero point to the population average (Grand Mean) of internationalization. The intercept therefore represents a club with 'typical' diversity for the league.

    -   Q: Does having a *higher proportion of foreign players than the average club* increase market value?

-   **Height**: Height is a static physical trait. The market values a 195cm defender because he is objectively tall, not because he is taller than his teammates. Grand mean centering allows us to interpret the coefficient as "the premium for being 1cm taller than the average soccer player. Keeps the intercept interpretable as the value of an "average height" player.

    -   Q: Does being taller than the average player in the league increase market value?

### **Year centering**

-   **UEFA** Coefficient = UEFA coefficients suffer from "grade inflation" and formula changes over decades; a raw score of 15.0 in 2013 is not equivalent to 15.0 in 2024. Centering by year isolates relative prestige—how dominant the club was *that specific season* compared to its peers. This removes the confounding effect of temporal trends in the coefficient system itself.

    -   Q: Does playing for a club that is currently outperforming the European average boost a player's value?

### **Year-Club mean centering (relative to teammates)**

-   **minutes_played_c**: Market value is driven by a player's status within their team. This metric isolates whether a player is a "Starter" (positive value) or a "Bench Player" (negative value) relative to their specific squad in that specific season. It removes the noise of "teams that rotate more" vs "teams that don't," focusing purely on the player's hierarchy in the squad.

    -   Q: Does being a key player (playing more than the average teammate) increase market value?

### **Position centering:**

-   **Goals, Assists:** A goal scored by a striker is expected; a goal scored by a defender is rare and highly valuable. If we leave `goals` and `assists` as raw counts, the model treats 1 goal the same for everyone, potentially undervaluing defensive contributions. We will group by position as performance expectations vary entirely by role. When we center within the position group, we transform the variable from a raw count into a "performance above expectation" metric. This interprets as the effect of scoring above position expectations.

    -   Q: Does scoring more goals than the average player in their position increase market value?" (This distinguishes a goal-scoring defender from an underperforming striker).

### **Non-centered fixed effects:**

*Performance Metrics*

-   **Red and Yellow card:** These variables have a meaningful 0 interpretation, and since our outcome is log transformed, these uncentered variables allow us to make percentage interpretations (i.e., one additional red card is associated with beta% change in market value).

    -   Q: What is the direct market penalty for receiving one additional red/yellow card?

-   **Position**: Reference group will be attackers.

    -   Q: How much less is a defender valued compared to an attacker, holding performance constant?

-   **Region**: Reference group European & Central Asia.

    -   Q: What is the market premium for players in the 'Europe' region compared to South America?

-   **Citizen**: Reference group is non-citizen of the country where the club is registered. The coefficient represents the simple difference in means between the two groups. A coefficient of +0.10 means "Citizens are valued 10% higher than Non-Citizens, all else held equal."

    -   Q: Is there a homegrown premium' (or discount) for players who hold citizenship in the club's country compared to foreign players?

Random Intercept and slope

-   Players: Yearly market values are nested within players.

    -   Q: After controlling for goals and age, how much variance in value is simply due to who the player is (marketing value/reputation)?

-   Club: Players are cross-classified within clubs.

    -   Q: How much of the variation in value is due to the baseline prestige of the club they play for?

-   Year: Without this random slope, the model assumes all clubs grow at the same rate. By including it, we allow for "Incubator Clubs" (steep growth) and "Stagnant Clubs" (flat/declining growth).

    -   Q: Does the rate of growth in player market value differ significantly between clubs?

\

**Why cross-classified model?**

Players can change teams over their span of their careers meaning that their player context can change. They may be a star player in one team and a bench player on a different team that they switch to. In a cross-classified model, the player is the constant unit (i), but the club (j) changes over time (t). If player (I) plays for club (j1) in year (t0) but switches to club (j2) in year (t1) the fixed effect in the model, such as minutes played, updates the players status, as opposed to having static or average predictors. Club attributes, such as foreign players and UEFA scores, ensures that when player move clubs, the model swaps the values of X_j1, for X_j2. Player’s characteristics, for instance age and height, move with the player across time.

**Assumptions about the data:**

We log transform market value to 2024 USD because financial data follows a pareto distribution rather than a normal distribution. If we keep the outcome on the raw scale, our residuals will be extremely large for high value players, and small for low value players violating the assumption of homoscedasticity. Log transforming the outcome compresses the extreme outliers by pulling the long right tail to try and make the distribution normal. We handle exchange rate fluctuation by using historical exchange rates to convert our values to USD.We also adjust market value to 2024 USD because 10 million in 2013 is not the same as in 2024 due to inflation, and by keeping the raw values we would lose real economic value. Therefore, this adjustment allows us to compare real economic power of a player in 2013 versus 2014 without the noise of currency fluctuation.

We assume that variables such as height and positions are time invariant for the duration of this study. At the club level, we assume mutually exclusiveness in that players can only belong to one club in a given year. Although a player could have switched clubs mid-season, we keep this constant for model simplicity. We assume that the relationship between age and log-market value is non-linear and parabolic so we introduce a quadratic term on age. Because we excluded observations with missing data (listwise deletion), we assume the data are Missing Completely at Random (MCAR) or that the proportion of missing cases (\~2%) is negligible and does not introduce systematic bias.

| **Variable Type** | **Variable** | Centering Strategy/ Form Factor | Justification |
|------------------|------------------|------------------|-------------------|
| Time | Season Year | Scale-Centered (+ Quadratic) | Centered at 2020, pandemic |
| Contextual, Club | UEFA Coefficient | Year-Mean Centering | Isolates relative prestige; removes historical inflation |
|  | Foreign Players % | Grand-Mean Centering | Moves baseline to "Average Club" (since 0% is unlikely) |
| Player Characteristics | Age | Grand-Mean (+ Quadratic) | Captures absolute career stage (Prospect $\to$ Prime $\to$ Veteran) |
|  | Height | Grand-Mean | Captures premium for absolute physical size |
|  | Position | Reference group, attacker | Attackers tend to be valued more on average |
| Player Demographic | Citizen of Club Country | None (Raw binary 0/1) | Keeps "Non-Citizen" as clear reference group. Separates individual status from club composition (`foreign_players`) |
|  | Region of origin | Reference group, Europe | Control variable for where player is from |
| Player Performance | Goals | Position-Centering | Preserves natural 0 and marginal value of 1 goal interpretation |
|  | Assists | Position-Centering | Preserves natural 0 and marginal value of 1 assist interpretation |
|  | Red Card | raw | Meaningful 0 interpretation |
|  | Yellow Card | raw | Meaningful 0 interpretation |
|  | Average minutes played in season | Year-Mean Centering, z-score | Isolates whether player is a starter or bench player |

: Data Structure for fixed effects, except year as it is a random slope

\

```{r}
#| echo: false
#| label: fig-cross-class
#| fig-cap: "Cross-Classified Data Structure: Players move between clubs over time."
#| fig-height: 4


# 1. Define the data INSIDE the chunk so R knows it exists during Render
tibble(
  player_label = c(rep("Player A (One Club)", 4), 
                   rep("Player B (Transfer)", 4), 
                   rep("Player C (Journeyman)", 4)),
  season_year = rep(2020:2023, 3),
  club_label = c(
    "Real Madrid", "Real Madrid", "Real Madrid", "Real Madrid",
    "Dortmund", "Dortmund", "Man City", "Man City",
    "Ajax", "Juventus", "Juventus", "Bayern Munich"
  )
) |> # 2. Create the plot
  ggplot(aes(x = season_year, y = player_label)) +
  geom_tile(aes(fill = club_label), color = "white", 
            linewidth = 1.5, width = 0.9, height = 0.6) +
  geom_text(aes(label = club_label), color = "white", 
            fontface = "bold", size = 3) +
  scale_fill_viridis_d(option = "E", alpha=.8, begin = .8, end=.1) +
  scale_x_continuous(breaks = 2020:2023) +
  labs(
    x = "Season Year",
    y = NULL,
    fill = "Club Context"
  ) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10, color = "black"),
    axis.text.x = element_text(size = 10)
  )

```

\

```{r}

player_df <- dat |> 
  mutate(
    player_id = factor(player_id),
    club_id   = factor(club_id),
    # reference groups
    position = fct_relevel(position, "Attack"),
    Region = fct_relevel(Region,  "Europe & Central Asia"),
    # outcome on log scale
    usd_revenue_2024_log = log(usd_revenue_2024),
    # season year
    # scale years with center 2020 for pandomic shock
    season_year_c = as.numeric(scale(season_year, center = 2020)),
    # grand mean centering
    foreign_players = foreign_players - mean(foreign_players, na.rm=TRUE),
    age_c = Age - mean(Age, na.rm=TRUE),
    height_c = height_in_cm - mean(height_in_cm, na.rm=TRUE)
  ) |> 
  # year mean centering
  group_by(season_year_c) |> 
  mutate(uefa_coefficient = 
           uefa_coefficient - mean(uefa_coefficient, na.rm=TRUE), 
         minutes_played = as.numeric(scale(minutes_played))) |> 
  # position centering
  group_by(position) |> 
  mutate(goals = goals - mean(goals, na.rm=TRUE),
         assists = assists - mean(assists, na.rm=TRUE)) |> 
  ungroup()
```

-   After checking missing values, we saw no more than 108 missing cases for Region, 103 for citizen of club country, and we decide to drop these cases as there are few and we are left with 7,023 cases.

```{r}
# Which variable has missing value
player_df |> 
  reframe(across(everything(), ~sum(is.na(.)))) |>
  select_if(~ . > 0) |> 
  pivot_longer(everything(), names_to = "Variable", 
               values_to = "Missing Cases") |> 
  slice(1:5) |> 
  arrange(desc(`Missing Cases`)) |> 
  kable()
  

clean_player_df <- player_df |> drop_na()
```

-   After examining the distribution of log market values across clubs, it is evident that the boxplots differ in both central tendency and variability. This indicates that club is a meaningful factor in explaining variation in market values.

```{r}
clean_player_df |>
  mutate(Club = fct_reorder(factor(Club), usd_revenue_2024_log, median)) |>
  ggplot(aes(x=Club, y=usd_revenue_2024_log, group=Club)) +
  stat_slabinterval(side = "both")  +                     
  theme(axis.text.x=element_text(angle = -60, hjust = 0)) +
  labs(title = "Distribution of log revenue by club",
       y="Log revenue (2024 USD)")


```

#### Linearity

In the overall spaghetti plot group by play_id, the lines vary widely across players, showing considerable between-player heterogeneity in log market value. Within most players, the slopes appear mostly flat, but some of them change up and down dramatically.

In the overall spaghetti plot group by club_id, they are consistently show differences.This means between-club differences in market values are large, supporting the idea that club-level random intercepts are necessary in your multilevel model.

Besides,the clubs’ average values change over time, indicating slopes might vary across clubs.

The loess smoother plot shows a non-linear averaged trend (curvature) in log-transformed market value (usd_revenue_2024_log) over centered season year (season_year_c). We can clearly see a peak at about season year 6 to 7, and the shape of the loess line is a wave pattern. Thus, we can try to add polynominal terms in our model to better explain the data.

```{r}


ggplot(clean_player_df) +
  geom_line(aes(x = season_year_c, y = usd_revenue_2024_log, 
              group = player_id),color = "black", alpha = 0.15)


ggplot(clean_player_df) +
  geom_line(aes(x = season_year_c, y = usd_revenue_2024_log, 
              group = club_id,color = factor(clean_player_df$club_id)),
              alpha = 0.5)


ggplot(clean_player_df, aes(x = season_year_c, 
                            y = usd_revenue_2024_log)) +
  
  # LOESS smoother (overall trend)
  geom_smooth(method = "loess",
              color = "lightblue", size = 1.2, se = FALSE) +
  labs(x = "Centered Season Year",
       y = "Log Revenue",
       title = "Spaghetti Plot with LOESS Trend") +
  theme_minimal(base_size = 12)



```

\

# Modeling Process Plan

-   Model 1: Random Intercepts only to examine ICC

-   Testing whether a random slope on year is needed.

    -   Model 2.1: Random intercepts + fixed effect of centered season_year

    -   Model 2.2: Adding quadratic terms for season year

    -   Model 2.3: Adding cubic term for season year

    -   Model 2.4 Adding random slope of season_year_c

    -   Model 2.5: Adding quadratic term of season_year as a random slope

-   Adding level fixed effects

    -   Establish whether age quadratic term is needed

        -   Model 3.1: Adding player covariates {Height + Position + Citizen Status + Region + Age + Goals + Assists + Red Cards + Yellow cards + Minutes played + UEFA coefficients + Foreign Players}

        -   Model 3.2: Adding the quadratic term for age

-   Compare model fit

-   Model diagnostic

-   Choose final model

\

### Model 1, intercepts only

```{r}

#Using bobyqa to handle the complex random effects structure and avoid gradient warnings
strict_control <- lmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 200000))

Model_1 <-  lmer(
  usd_revenue_2024_log ~ 
    # random effects
      (1|player_id) + (1|club_id), ,  
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_1)
```

-   We can see that player_id accounts for about 75% of market value while club_id accounts for 3%.

\

### Model 2.1: Adding fixed effect of year

```{r}
Model_2.1 <-  lmer(
  usd_revenue_2024_log ~ 
    # first level fixed effect
    season_year_c +
    # random effects
      (1|player_id) + (1|club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_2.1)
```

\

### Model 2.2 Adding quadratic term

-   Notice that the linear relationship of year is no longer significant.

```{r}
Model_2.2 <-  lmer(
  usd_revenue_2024_log ~ 
    # first level fixed effect
    # Use poly() with raw=TRUE 
    poly(season_year_c, 2, raw = TRUE) + 
    # random effects
      (1|player_id) + (1|club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_2.2)
```

\

#### Compare Model 2.1 and 2.2 to determine if the quadratic term is needed.

-   We see that the LRT test suggests that the full model provides a significantly better fit than the reduce model, the quadratic term is needed for this data.

```{r}
anova(Model_2.1, Model_2.2, test = "LRT") |> gt()
```

\

### Model 2.3 Adding cubic term

-   Now linear year trends is significant

```{r}

Model_2.3 <-  lmer(
  usd_revenue_2024_log ~ 
    # first level fixed effect
    # Use poly() with raw=TRUE 
    poly(season_year_c, 3, raw = TRUE) + 
    # random effects
      (1|player_id) + (1|club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_2.3)
```

\

#### Compare Model 2.2 and 2.3 to determine if the quadratic term is needed.

-   We see that the LRT test suggests that the full model provides a significantly better fit than the reduce model, the cubic term is needed for this data. After adding cubic terms, it yields a better fit. It cannot only explain the peak of the overall trajectory, but can also explain the up and down wave pattern over time of the trajectory.

```{r}
anova(Model_2.2, Model_2.3, test = "LRT") |> gt()
```

\

### Model 2.4, Random slope

-   To estimate a random slope `(1 + year_centered | player_id)`, the model needs multiple data points for *every* player to calculate their individual trajectory. But many players only appear in our data for 1 or 2 years, and the model struggles to distinguish the "player slope" from the "residual error," causing these convergence issues. Clubs, however, have many players and many years of historical market value data so a random slope for clubs is more robust to fit.

```{r}
Model_2.4 <-  lmer(
  usd_revenue_2024_log ~ 
    ## first level fixed effect
    # Use poly() with raw=TRUE 
    poly(season_year_c, 3, raw = TRUE) + 
    # random effects
      (1 | player_id) + (poly(season_year_c, 1, raw = TRUE) | club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_2.4)
```

### Model 2.5, Random slope + quadratic term

```{r}

Model_2.5 <-  lmer(
  usd_revenue_2024_log ~ 
    # first level fixed effect
    # Use poly() with raw=TRUE 
    poly(season_year_c, 3, raw = TRUE) +  
    # random effects
      (1 | player_id) + (poly(season_year_c, 2, raw = TRUE) | club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_2.5)
```

```{r}
anova(Model_2.4, Model_2.5, test = "LRT") |> gt()
```

\

### Model 2.6, Random slope + quadratic term + cubic

```{r}
Model_2.6 <-  lmer(
  usd_revenue_2024_log ~ 
    # first level fixed effect
    # Use poly() with raw=TRUE 
    poly(season_year_c, 3, raw = TRUE) +  
    # random effects
      (1 | player_id) + (poly(season_year_c, 3, raw = TRUE)| club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_2.6)
```

```{r}
anova(Model_2.3, Model_2.5, Model_2.6, test = "LRT") |> gt()
```

-   We see that Model 2.6 with the cubic slope failed to converge as we are asking the model to estimate too many parameters for very few clubs (21 in total). Additionally, it is marginally better than the Model 2.5, but with a higher BIC value. We keep Model 2.5 for simplicity.

### Determine if the random slope on season year is needed

```{r}
# LRT test for 2 random slopes
calculate_lrt_mixture <- function(reduced_model, full_model) {
  
  # 1. Extract Log-Likelihoods
  ll_reduced = as.numeric(logLik(reduced_model))
  ll_full = as.numeric(logLik(full_model))
  
  # 2. Calculate LRT Statistic
  lrt_stat = -2 * (ll_reduced - ll_full)
  
  # 3. Calculate Difference in Parameters (df)
  df_reduced = attr(logLik(reduced_model), "df")
  df_full = attr(logLik(full_model), "df")
  df_diff = df_full - df_reduced
  
  # 4. Calculate P-Value based on the difference
  
  if (df_diff == 3) {
    # --- NEW BLOCK FOR QUADRATIC SLOPE ---
    # Going from 2 Random Effects (Int+Lin) -> 3 Random Effects (Int+Lin+Quad)
    # We add 1 variance + 2 covariances = 3 parameters.
    # The mixture is 0.5 * ChiSq(2) + 0.5 * ChiSq(3)
    p_val = 0.5 * pchisq(lrt_stat, df = 2, lower.tail = FALSE) + 
             0.5 * pchisq(lrt_stat, df = 3, lower.tail = FALSE)
    
    msg = "Mixture of ChiSq(2) and ChiSq(3)"

  } else if (df_diff == 2) {
    # Going from 1 Random Effect (Int) -> 2 Random Effects (Int+Lin)
    # Adds 1 variance + 1 covariance = 2 parameters
    p_val = 0.5 * pchisq(lrt_stat, df = 1, lower.tail = FALSE) + 
             0.5 * pchisq(lrt_stat, df = 2, lower.tail = FALSE)
    
    msg = "Mixture of ChiSq(1) and ChiSq(2)"
    
  } else if (df_diff == 1) {
    # Adding a single parameter (e.g. just a variance, no covariance)
    p_val = 0.5 * pchisq(lrt_stat, df = 0, lower.tail = FALSE) + 
             0.5 * pchisq(lrt_stat, df = 1, lower.tail = FALSE)
    
    msg = "Mixture of ChiSq(0) and ChiSq(1)"
    
  } else {
    # Fallback for standard fixed effects comparisons
    p_val = pchisq(lrt_stat, df = df_diff, lower.tail = FALSE)
    msg = paste0("Standard ChiSq(", df_diff, ")")
  }
  
  # Output
  cat("LRT Statistic:", round(lrt_stat, 4), "\n")
  cat("Parameter Diff:", df_diff, "\n")
  cat("Distribution:  ", msg, "\n")
  cat("P-Value:       ", format.pval(p_val, eps = .0001), "\n")
}

# Run the comparison
calculate_lrt_mixture(Model_2.3, Model_2.5)


```

-   To determine if the rate of market value growth varies significantly across clubs, we compared the random-intercept only model with season_year_c linear and quadratic fixed effects to the model with a random linear and quadratic slopes using a Likelihood Ratio Test (LRT). The Null hypothesis tests whether the variance parameter on the boundary of the parameter space ($H_0: \sigma^2_{slope} = 0$), we utilized a helper function to evaluate significance against a correct 50:50 mixture distribution of $.5 \times \chi^2(1) + .5 \times \chi^2(2)$. The test yielded a significant results ($\chi^2=97.13, p<.001$) leading us to reject the Null. This provides strong evidence that clubs do not follow parallel trends; rather, there is significant heterogeneity in club trajectories, justifying the inclusion of the random slope.

\

### Model 3.1: Adding fixed effects at level 1 with age as a linear relationship

-   Age centered is not even significant. Same for red\\yellow cards.\

```{r}
Model_3.1 <-  lmer(
  usd_revenue_2024_log ~ 
    # first level fixed effect
    # Use poly() with raw=TRUE 
    poly(season_year_c, 3, raw = TRUE) + 
    # player fixed effects
    height_c + position + citizen_of_club_country + Region +
    age_c + 
    goals + assists + red_cards + yellow_cards + minutes_played +
    # random effects
    (1 | player_id) + (poly(season_year_c, 2, raw = TRUE) | club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_3.1)
```

### Model 3.2. Adding age quadratic term

```{r}

Model_3.2 <-  lmer(
  usd_revenue_2024_log ~ 
    # first level fixed effect
    # Use poly() with raw=TRUE 
    poly(season_year_c, 3, raw = TRUE) + 
    # player fixed effects
    height_c + position + citizen_of_club_country + Region +
    poly(age_c, 2, raw = TRUE) +
    goals + assists + red_cards + yellow_cards + minutes_played +
    # club fixed effects
    uefa_coefficient + foreign_players + 
    # random effects
    (1 | player_id) + (poly(season_year_c, 2, raw = TRUE) | club_id), 
  data = clean_player_df, REML = TRUE,   control = strict_control)

summ(digits = 4,  Model_3.2)
```

-   It seems that we need the quadratic term for age in the model.

```{r}
anova(Model_3.1, Model_3.2) |> gt()

```

\

```{r}
export_summs(Model_2.5, Model_3.2,
              digits=3, r.squared=FALSE, 
             model.names = c("Model_2.5", "Model_3.2")) 


final_model <- Model_3.2
```

\

### Model Diagnostics

-   By comparing the QQ-line and QQ-plot (both residuals and random effects) we can find that both model plots look good on normality, except some outliers. By comparing the residual-fitted value plot, we can find that both model show the same spread pattern, the dots are evenly spread across zero-mean line up and down.

-   By comparing EBLUPs plots, we can see that most of the points lie deviate from the diagonal at both tails, and the lower side is more downward, suggesting moderate departures from normality in the random slopes and random intercepts, which indicates possible skewness and outliers.Thus, we cannot say the normal random effects assumption in the multilevel model are met.

```{r}


# slopes
qqnorm(ranef(final_model)$player_id[,1]) # group factor
qqline(ranef(final_model)$player_id[,1])


qqnorm(ranef(final_model)$club_id[,1]) # group factor
qqline(ranef(final_model)$club_id[,1])

qqnorm(ranef(final_model)$club_id[,2]) # group factor
qqline(ranef(final_model)$club_id[,2])

```

-   The red regression line slopes slightly downward, but only gently. That is, clubs with higher baseline log market value tend to have slightly more negative or flatter slopes. Clubs with lower baseline market value tend to have slightly steeper slopes.

-   Most residuals lie between about −1 and +1, and fitted log revenues between 14 and 18. There is no strong curved pattern or obvious systematic relationship between residual and fitted values. So the residual reasonably consistent with the linearity assumption. Besides, there is no funnel shape, so the residuals are homogeneous. However, we can still see some outliers exist.

```{r}
# 1. Extract the random effects for clubs
re <- ranef(final_model)$club_id

# 2. Rename columns for clarity (poly names are messy)
# Note: Ensure your model actually has 3 REs. If not, adjust indices.
colnames(re) <- c("Intercept", "Linear_Slope", "Quad_Slope")

# 3. Create a Scatterplot Matrix (Pairs Plot)
# This plots every random effect against every other random effect
pairs(re, 
      main = "Correlation between Random Effects",
      pch = 19, 
      col = "dodgerblue",
      lower.panel = panel.smooth, # Adds trend lines automatically
      upper.panel = NULL)         # Hides duplicate upper half
```

```{r}
#| fig-width: 7
#| fig-height: 10
#| 
check_model(final_model)
```

\newpage

## Final Model Specification

We specify a **cross-classified multilevel model** (Model 3.2) to partition the variance in market value across observations ($t$), players ($i$), and clubs ($j$). Crucially, because club attributes like internationalization and UEFA ranking vary year-to-year, they are modeled as **Level 1 (time-varying)** predictors rather than static Level 2 traits.

**Level 1: Observation Model (Time-Varying Effects)**

At the observation level, the log-transformed market value is modeled as a function of the player-club intercept ($\pi_{0ij}$), algebraic time trends, biological aging, and dynamic contextual metrics:

$$
\begin{aligned}
\log(\text{Revenue})_{tij} &= \pi_{0ij} \\
&+ \pi_{1j}(\text{Year}_{t}) + \pi_{2j}(\text{Year}^2_{t}) + \pi_{3}(\text{Year}^3_{t}) \\
&+ \pi_{4}(\text{Age}_{it} - \bar{A}) + \pi_{5}(\text{Age}_{it} - \bar{A})^2 \\
&+ \pi_{6}(\text{Mins}_{Z,it}) \\
&+ \pi_{7}(\text{Foreign}_{tj}) + \pi_{8}(\text{UEFA}_{tj}) \\
&+ \mathbf{\Pi}_{perf} + \epsilon_{tij}
\end{aligned}
$$

*Note: The subscripts ${tj}$ for Foreign and UEFA denote that these club attributes vary by time $t$.*

**Level 2a: Player Model (Time-Invariant Traits)**

The player-club intercept ($\pi_{0ij}$) is decomposed into a fixed component based on static player characteristics and a random player effect ($u_{0i}$):

$$
\pi_{0ij} = \beta_{0j} + \gamma_{1}(\text{Height}_{i}) + \gamma_{2}(\text{Citizen}_{ij}) + \mathbf{\Gamma}_{demog} + u_{0i}
$$

**Level 2b: Club Model (Contextual Trajectories)**

The club baseline ($\beta_{0j}$) and the time-varying slope parameters ($\pi_{1j}, \pi_{2j}$) include random effects to allow for heterogeneity in both growth velocity and trajectory curvature:

$$
\begin{aligned}
\beta_{0j} &= \delta_{00} + v_{0j} \quad &\text{(Baseline Prestige)} \\
\pi_{1j} &= \delta_{10} + v_{1j} \quad &\text{(Linear Growth Deviation)} \\
\pi_{2j} &= \delta_{20} + v_{2j} \quad &\text{(Quadratic Curvature Deviation)} \\
\pi_{3} &= \delta_{30} \quad &\text{(Fixed Cubic Trend)}
\end{aligned}
$$

**Variance-Covariance Structure**

$$
u_{0i} \sim N(0, \sigma^2_{u}) \quad , \quad 
\begin{pmatrix} v_{0j} \\ v_{1j} \\ v_{2j} \end{pmatrix} \sim N \left( \mathbf{0}, \begin{pmatrix} \sigma^2_{Int} & \sigma_{01} & \sigma_{02} \\ \sigma_{01} & \sigma^2_{Lin} & \sigma_{12} \\ \sigma_{02} & \sigma_{12} & \sigma^2_{Quad} \end{pmatrix} \right)
$$

------------------------------------------------------------------------

## Interpretation of Results

**Temporal Dynamics: The "Great Pause"**
Using raw polynomials centered at 2020 ($t=0$), our model mathematically pinpoints the moment the market froze. The **linear coefficient at the 2020 baseline is effectively zero** ($\pi_1 = 0.00, p = 0.93$), indicating that at the exact onset of the pandemic, the relentless inflation of the previous decade came to a complete halt. The significant negative quadratic ($\pi_2 = -0.27$) and cubic ($\pi_3 = -0.12$) terms confirm the transition from a pre-2019 "Boom" into a structural correction.

**Club Context and Diversity**
While we hypothesized a "Diversity Premium" for clubs with higher international representation, the fixed effect for `foreign_players` was positive but **not statistically significant** ($\pi_7 = 0.10, p = 0.32$) in the final model. Similarly, the `uefa_coefficient` showed no significant effect ($p = 0.70$). This suggests that once we account for the massive heterogeneity in **Club Random Effects** (Baseline Prestige and Growth Trajectories) and individual player reputation ($ICC \approx 0.78$), specific club-level metrics add little unique explanatory power.

**Squad Status and Homegrown Effects**
Squad status remains the dominant behavioral driver. The standardized coefficient for minutes played ($\pi_6 = 0.35$) indicates a **42% valuation premium** ($e^{0.35}-1$) for key starters (+1 SD) compared to average players. Conversely, we observe a robust **"Domestic Discount,"** where homegrown citizens are valued approximately **17% lower** ($\gamma_2 = -0.19$) than comparable foreign imports.

```{r}

# average age
mean_age <- clean_player_df |> reframe(mean(Age, na.rm=TRUE)) |> pull()

# Simulate from the estimated sampling distribution
# "-b1/(2*b2)" Report with confidence interval using delta method

sims <- sim(final_model, n.sims = 1000) 

peak_sims <- sims@fixef |> 
  as.data.frame() |> 
  tibble() |> 
  reframe( mean(
    (`poly(age_c, 2, raw = TRUE)1` * -1) / 
      (2 * `poly(age_c, 2, raw = TRUE)2`) )
    ) |> 
  pull()



# Empirical CI
cat("Peak Age = ", round(mean_age + peak_sims, 2))

# The delta method uses calculus (specifically, Taylor series approximation) 
## to propagate uncertainty through nonlinear transformations.
```

\newpage

## Visualization of Marginal Effects

### **Q1: Market Evolution**

How does market value change annually compared to the 2020 baseline?

-   The visualization answers **Q1** by displaying the market's trajectory in calendar time. The curve peaks just prior to the **2020 baseline**, confirming that the rapid inflation of the late 2010s halted abruptly at the onset of the pandemic. Post-2020, the curve flattens and dips, illustrating the lasting structural correction in player valuations.

```{r}
# 1. Calculate Scaling Parameters from Data
# We need the exact SD used in the model to reverse the math
year_sd <- sd(clean_player_df$season_year, na.rm = TRUE)
year_center <- 2020 # You explicitly set center=2020

# 2. Generate Predictions (on Z-score scale)
pred_trend <- ggpredict(Model_3.2, terms = "season_year_c [all]") 

# 3. Back-Transform the X-axis
plot_data <- pred_trend %>%
  as.data.frame() %>%
  mutate(
    # Formula: Real_Year = (Z_Score * SD) + Center
    year_real = (x * year_sd) + year_center
  )

# 4. Plot using Real Years
ggplot(plot_data, aes(x = year_real, y = predicted)) +
  
  # Trend Line and Ribbon
  geom_line(color = "#440154", linewidth = 1.5) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "#440154", alpha = 0.2) +
  
  # Reference Line for 2020
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50") +
  annotate("text", x = 2020.2, y = max(plot_data$predicted), 
           label = "Pandemic Start", hjust = 0, vjust = -2,
           size = 4, color = "gray40") +
  
  # Scale X-axis to show clean years
  scale_x_continuous(breaks = seq(2013, 2024, 2)) +
  
  labs(
    title = "A. Market Evolution (Cubic)",
    subtitle = "Predicted values back-transformed to calendar years",
    x = "Season Year", 
    y = "Log Market Value"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

\

\

Below we visualize the predicted market value evolution for individual players during their active seasons, holding club and global market trends constant. The plot confirms that top-tier players follow a fundamentally different valuation logic than the general population: while the average player (gray lines) exhibits a relatively flat trajectory centered around the league mean, "Superstars" (colored lines) show massive vertical shifts and often steeper acceleration phases that outpace biological maturity. This "fanning out" of curves, particularly post-2020, mathematically illustrates the "Superstar Effect", where the gap between elite outliers and the field widens over time, and suggests that individual brand value can serve as a buffer against sector-wide downturns like the pandemic shock.

```{r}
#| fig-width: 8
#| fig-height: 10
# 1. CREATE DATA FOR PLOTTING (Simple Centering, NOT Z-score)
# This ensures -7 = 2013 and +4 = 2024
plot_data_prep <- clean_player_df %>%
  mutate(season_year_c = season_year - 2020) %>%
  # Ensure other variables match what the formula expects
  mutate(
    minutes_played = as.numeric(scale(minutes_played)), # Z-score minutes
    age_c = Age - mean(Age, na.rm=TRUE) # Simple centering age
  )

# 2. REFIT MODEL ON THIS SPECIFIC DATA
# We must re-fit so the coefficients align with the -7 to +4 scale
Plot_Model <- lmer(
  usd_revenue_2024_log ~ 
    poly(season_year_c, 3, raw = TRUE) + # Use raw for plotting model
    height_c + position + citizen_of_club_country + Region +
    poly(age_c, 2, raw = TRUE) + 
    goals + assists + red_cards + yellow_cards + minutes_played +
    # Note: Model 3.2 structure (No Foreign/UEFA fixed effects)
    (1 | player_id) + (poly(season_year_c, 2, raw = TRUE) | club_id),
  data = plot_data_prep, 
  control = lmerControl(optimizer = "bobyqa")
)

# 3. Create Metadata Dictionary
club_meta <- plot_data_prep %>%
  mutate(club_id = as.character(club_id)) %>%
  group_by(club_id) %>%
  summarise(Club = first(Club)) %>% 
  ungroup()

# 4. Generate Predictions (Full Range)
# using [all] on the REFITTED model will now span -7 to +4
club_pred <- ggpredict(Plot_Model, 
                       terms = c("season_year_c [all]", "club_id"), 
                       type = "random") 

# 5. Process Data
final_plot_data <- club_pred %>%
  as.data.frame() %>%
  mutate(group = as.character(group)) %>% 
  left_join(club_meta, by = c("group" = "club_id")) %>%
  group_by(Club) %>%
  mutate(max_val = max(predicted)) %>%
  ungroup() %>%
  mutate(Club = reorder(Club, -max_val))

# 6. Plot
ggplot(final_plot_data, aes(x = x + 2020, y = predicted)) +
  
  # Pandemic Line
  geom_vline(xintercept = 2020, linetype = "dashed", 
             color = "red", alpha=0.5) +
  
  # Trajectories
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "#440154", alpha = 0.1) +
  geom_line(color = "#440154", linewidth = 1.2) +
  
  # Facets
  facet_wrap(~Club, ncol = 5) + 
  
  # Ensure axis covers full range
  scale_x_continuous(breaks = seq(2013, 2024, 2), 
                     limits = c(2013, 2024)) +
  
  labs(
    title = "Club Trajectories Through the Pandemic (2013-2024)",
    subtitle = "Predicted Log Market Value (Refitted Model)",
    x = "Season Year",
    y = "Log Market Value (2024 USD)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Q2: Biological Career Arc

How does market value change as a player moves through their biological career arc (Prospect $\to$ Prime $\to$ Veteran)?

-   Below we visualize the biological constraints on valuation. The curve rises steeply as players transition from 'Prospects' (Age 17–23), flattens into a 'Prime' plateau between ages 26 and 28, and declines steadily thereafter. The model predicts that a player's value drops by approximately 1.5 log points (from a peak of 17.0 to 15.5) as they age from 27 to 35, reflecting the market's heavy discount on older assets with limited resale potential.

```{r}
# 1. Get Mean Age to reverse the centering
# We need the exact mean used in the dataset
mean_age <- mean(clean_player_df$Age, na.rm = TRUE)

# 2. Generate Predictions
pred_age <- ggpredict(Model_3.2, terms = "age_c [all]") 

# 3. Back-Transform X-axis
plot_data_age <- pred_age %>%
  as.data.frame() %>%
  mutate(
    # Formula: Real_Age = Centered_Age + Mean_Age
    age_real = x + mean_age
  )

# 4. Plot with Real Ages
ggplot(plot_data_age, aes(x = age_real, y = predicted)) +
  
  # Trend Line and Ribbon
  geom_line(color = "#3b528b", linewidth = 1.5) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "#3b528b", alpha = 0.2) +
  
  # Add Peak Reference Line (Visual aid)
  geom_vline(xintercept = 27, linetype = "dashed", color = "gray50") +
  annotate("text", x = 27.5, y = min(plot_data_age$predicted) + 1, 
           label = "Peak ~27", hjust = 0, color = "gray40", size = 3.5) +
  
  # formatting
  scale_x_continuous(breaks = seq(16, 40, 4)) + # Show ages 16, 20, 24...
  
  labs(
    title = "B. The Career Arc",
    subtitle = "Predicted Log Market Value by Biological Age",
    x = "Player Age (Years)", 
    y = "Log Market Value"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

### Q3: Squad Status

Does being a key player (playing more than the average teammate) increase market value?

-   Figure 1C visualizes the massive premium placed on utilization. The steep linear slope confirms that 'Squad Status' is a primary driver of valuation. Moving from a rotational role (−1 SD) to a key starter role (+1 SD) is associated with an approximate doubling of market value (+101%), indicating that the market heavily discounts latent talent that cannot crack the starting XI.

```{r}
# Generate predictions for Minutes Played
pred_status <- ggpredict(Model_3.2, terms = "minutes_played [all]") 

plot(pred_status) +
  geom_line(color = "#21918c", linewidth = 1.5) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#21918c", alpha = 0.2) +
  labs(
    title = "C. Squad Status",
    x = "Minutes Played (Z-Score)", 
    y = "Log Market Value"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

\

\

### **Contextual Premiums and Discounts (Q5-Q7)** Figure 1D visualizes the valuation gaps across citizenship, position, and region.

-   **The "Domestic Discount" (Q5):** The coefficient for "Homegrown (Citizen)" falls strictly to the left of zero, confirming that domestic players are valued significantly lower than comparable foreign imports ($\beta=−0.19$), potentially reflecting the recruitment premium required to attract international talent.

-   **Positional Hierarchy (Q6):** Attackers (the baseline) command the highest market value. All other positions show significant negative effects, with Goalkeepers facing the steepest penalty ($\beta=−1.06$), suggesting the market disproportionately rewards offensive output.

-   **The "Export Brand" Effect (Q7):** While most regional coefficients overlap with zero (indicating no difference from the European baseline), **Latin America** stands out as the only region with a positive premium ($\beta=0.19$). This suggests that South American players function as a "premium export brand" in the global market, distinct from other non-European origins.

```{r}
plot_summs(Model_3.2, 
           coefs = c("Homegrown (Citizen)" = "citizen_of_club_countryYes",
                     "Defender" = "positionDefender",
                     "Midfielder" = "positionMidfield",
                     "Goalkeeper" = "positionGoalkeeper",
                     "Latin America" = "RegionLatin America & Caribbean",
                     "Asia & Pacific" = "RegionEast Asia & Pacific",
                     "North America" = "RegionNorth America",
                     "Middle East & N. Africa" = "RegionMiddle East & North Africa",
                     "Sub-Saharan Africa" = "RegionSub-Saharan Africa"),
           colors = "#440154", # Viridis Purple
           point.shape = 18) + 
  
  labs(title = "D. Relative Valuation (Q5, Q6, Q7)",
       subtitle = "Discounts (Left) vs. Premiums (Right) relative to baseline",
       x = "Effect on Log Market Value") +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

\

Linear Trend Deviation (Center Panel): Clubs to the right (positive values) experienced faster overall value growth across the decade compared to the market average. Curvature Deviation (Right Panel): Positive values indicate a 'Convex' shape (accelerating growth or strong recovery), while negative values indicate a 'Concave' shape (diminishing returns or plateauing).

### Club-Level Heterogeneity

Figure below displays the estimated club random effects from our cross-classified model. Substantial heterogeneity exists in baseline valuations, temporal trends, and curvature.

-   **Baseline Prestige (Panel A):** As expected, elite 'Super Clubs' such as **Manchester City** and **Bayern Munich** show massive positive intercepts, indicating that their players command a premium of \~30% purely due to the club's brand.

-   **Linear Trend Deviation (Panel B):** Clubs to the right (positive values) experienced **faster overall value growth** across the decade compared to the market average. For example, **Arsenal** and **Liverpool** show strong positive deviations, confirming their role as high-growth environments.

-   **Curvature Deviation (Panel C):** This panel reveals the *shape* of the trajectory. Positive values indicate a **'Convex'** shape (accelerating growth or resilient recovery), while negative values indicate a **'Concave'** shape (diminishing returns or plateauing), highlighting the different ways clubs navigated the market cycle.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 10
#| # 1. Extract Random Effects from Model 3.2
club_ranef <- ranef(Model_3.2, condVar = TRUE)$club_id
post_var <- attr(club_ranef, "postVar")

# 2. Metadata Dictionary
club_meta <- clean_player_df %>%
  mutate(club_id = as.character(club_id)) %>%
  group_by(club_id) %>%
  summarise(Club = first(Club)) %>% 
  ungroup()

# 3. Process Data (Robust for 3 Effects)
club_random_results <- club_ranef %>%
  as.data.frame() %>%
  rownames_to_column("club_id") %>%
  
  # Rename by position: 1=ID, 2=Int, 3=Linear, 4=Quadratic
  rename(Intercept = 2) %>%
  { if(ncol(.) > 2) rename(., Linear = 3) else . } %>%
  { if(ncol(.) > 3) rename(., Quadratic = 4) else . } %>%
  
  # Calculate CIs
  mutate(
    se_int = sqrt(post_var[1, 1, ]),
    se_lin = if("Linear" %in% names(.)) sqrt(post_var[2, 2, ]) else NA,
    se_quad = if("Quadratic" %in% names(.)) sqrt(post_var[3, 3, ]) else NA,
    
    Intercept_lower = Intercept - 1.96 * se_int,
    Intercept_upper = Intercept + 1.96 * se_int,
    
    Linear_lower = if("Linear" %in% names(.)) Linear - 1.96 * se_lin else NA,
    Linear_upper = if("Linear" %in% names(.)) Linear + 1.96 * se_lin else NA,
    
    Quadratic_lower = if("Quadratic" %in% names(.)) Quadratic - 1.96 * se_quad else NA,
    Quadratic_upper = if("Quadratic" %in% names(.)) Quadratic + 1.96 * se_quad else NA
  ) %>%
  
  # Note: We do NOT convert to percentage here because orthogonal polys are abstract units
  # We just plot the raw deviation values.
  
  left_join(club_meta, by = "club_id") %>%
  select(Club, Intercept, Linear, Quadratic, ends_with("lower"), ends_with("upper")) %>%
  pivot_longer(cols = -Club, names_to = "Key", values_to = "Value") %>%
  separate(Key, into = c("Type", "Stat"), sep = "_", fill = "right") %>%
  mutate(Stat = replace_na(Stat, "Estimate")) %>%
  pivot_wider(names_from = Stat, values_from = Value) %>%
  drop_na(Estimate)

# 4. Plot
ggplot(club_random_results, aes(x = reorder(Club, Estimate), y = Estimate, color = Type)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(ymin = lower, ymax = upper), size = 0.5) +
  
  facet_wrap(~Type, scales = "free_x", nrow = 1,
             labeller = as_labeller(c(
               "Intercept" = "Baseline Prestige", 
               "Linear" = "Linear Trend Deviation",
               "Quadratic" = "Curvature Deviation"))) +
  
  coord_flip() +
  scale_color_viridis_d(option = "D", end = 0.8) +
  labs(
    title = "Club-Level Trajectory Deviations",
    subtitle = "Estimates from Model 3.2 (Orthogonal Polynomials)",
    x = NULL, y = "Deviation from Market Average"
  ) +
  theme(legend.position = "none")

  
```

## The "Prodigy Premium": Pricing Future Potential

The largest positive random effects belong overwhelmingly to teenage phenoms, confirming that the market prices **potential** far higher than realized performance. **Lamine Yamal** exhibits the most extreme effect in the dataset: controlling for his age, position, and playing time, his market value is approximately **19.0 times higher** ($95\% \text{ CI}: [8.4, 43.1]$) than the model predicts. Similar "super-premiums" characterize other young stars like **Gavi** ($16.1\times$), **Florian Wirtz** ($15.5\times$), and **Jude Bellingham** ($11.8\times$).

These extreme deviations highlight a fundamental feature of football valuation: **Option Value**. Clubs pay premiums for young players not for their current statistical output (which is often modest compared to prime-age veterans), but because they are purchasing a "call option" on future superstar performance. The random intercept absorbs this unmeasured "star quality"—a combination of technical ceiling and marketability that scouts perceive but standard metrics cannot capture.

**The Barcelona Effect** There is a striking concentration of talent from FC Barcelona among the top outliers. Five of the top 11 premiums belong to Barcelona academy graduates or players: **Lamine Yamal** (1st), **Gavi** (2nd), **Pedri** (7th, $11.1\times$), **Ansu Fati** (10th, $9.5\times$), and **Pau Cubarsí** (11th, $9.0\times$). This suggests a potential **institutional multiplier**: the "La Masia" brand (Barcelona's academy) may systematically inflate the market's valuation of its graduates, or perhaps the club provides a unique platform that accelerates reputation faster than sporting output alone.

## The "Legacy Anomaly"

Interestingly, the list of outliers is not exclusively young. **Gianluigi Buffon** (8th, $10.7\times$) and **Cristiano Ronaldo** (18th, $7.2\times$) appear alongside the teenagers. This reflects a **"Legacy Premium."** Just as prodigies are priced on future potential, icons are priced on historical achievement and commercial gravity. Our model, which relies on current-season statistics and penalizes advanced age via the quadratic term, underestimates these players because it cannot see their career history or commercial value. The random intercept thus serves as a catch-all for **Reputation**, capturing both the promise of the future (Yamal) and the glory of the past (Ronaldo).

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 10
#| 
# extract random effects from the final model
player_ranef <- ranef(final_model, condVar = TRUE)$player_id 

player_random_results <- player_ranef |> 
  as.data.frame() |>
  rownames_to_column() |> 
  rename(player_id=1, Intercept=2) |> 
  # add club names from the dataframe
  left_join(
    clean_player_df |> 
      distinct(player_id, name)) |> 
  relocate(name) |> 
  select(-player_id) |> 
  tibble()

# Extract posterior variances (stored as an attribute)
post_var <- attr(player_ranef, "postVar")
se_intercept <- sqrt(post_var[1, 1, ])

# post_var is a 3D array: [effect, effect, club]
# post_var[1,1,] = variance of intercept for each club
# post_var[2,2,] = variance of slope for each club
# post_var[1,2,] = covariance between intercept and slope

# add standard errors to club_random_results
player_random_results <- player_random_results |> 
  mutate(se_intercept = se_intercept,
         # calculate 95% CI
         lower = Intercept - 1.96 * se_intercept,
         upper = Intercept + 1.96 * se_intercept
         ) |> 
  # convert to %
  mutate(across(where(is.double), ~ exp(.x)),
         name = fct_reorder(name, Intercept, max)) |> 
  select(-starts_with("se_")) 
  

# build catapiler visual of random effects
player_random_results |> 
  arrange(desc(Intercept)) |> 
  slice(1:50) |> 
  ggplot(aes(x = name, y = Intercept)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  geom_point(show.legend = FALSE) +
  geom_errorbar(aes(ymin=lower, ymax=upper), show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Player-Level Market Value Premium, Top 50",
    x = NULL,
    y = "Random Intercept (log market value deviation)",
    caption = 
      "Error bars show 95% confidence intervals. Dashed line = Player average.") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()) 
```

# Discussion

## Limitations

While our model demonstrates strong predictive power ($R^2 \approx 0.89$), several limitations warrant consideration:

1.  **Small Cluster Size:** We estimated a complex random effects structure (Intercept + Linear + Quadratic) for `Club`, despite our dataset containing only 21 unique clubs. While the Likelihood Ratio Test supported the inclusion of these slopes ($p < .001$), variance estimates derived from such a small number of clusters should be interpreted with caution.
2.  **Normality of Residuals:** Diagnostic plots indicated deviation from normality in the extreme tails of the distribution. While the large sample size ($N=7,023$) ensures robust estimation of fixed effects, future iterations could employ robust standard errors or a Gamma GLMM to better handle these extremes.
3.  **Endogeneity:** Predictors such as `minutes_played` are likely endogenous—since better players naturally accrue more playing time—and thus should be interpreted as associations reflecting "Squad Status" rather than purely causal effects.

## 6.4 Conclusion

Ultimately, this research highlights the necessity of moving beyond static, single-level analyses to capture the true complexity of the football transfer market. By rigorously modeling the cross-classified intersection of individual player trajectories and dynamic club contexts, we identified a **"Great Pause"** in 2020 where market growth hit zero, a strict biological valuation peak at age 27, and a massive 42% premium for starters. As the industry continues to professionalize, frameworks that account for the interplay of human capital, institutional prestige, and market cycles will remain indispensable for distinguishing genuine value from market noise.
