---
title: "Model Interpretation and Statistical Inference"
author: "Moneyballers"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: html
theme: lumen
toc: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(arrow)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(dplyr)
library(ggplot2)
library(emmeans)
library(performance)
library(jtools)
library(knitr)
library(kableExtra)
library(tidyr)

# Reading data with UEFA rankings
repo_path <- "/Users/namomac/Desktop/Moneyball_FC/"
clean_player_df <- read_parquet(paste0(repo_path, "inputs/player_df_with_uefa.parquet"))
```

## Model Estimation

```{r fit_final_model}
# Fitting final model with UEFA coefficient
m_final <- lmer(
  log(usd_revenue_2024) ~ 
    season_year + yellow_cards + goals + assists +
    position + age_cen + height_cen + 
    Region + citizen_of_club_country +
    uefa_coef_cen +  # Level 3: Club prestige
    (1 | club_id) + (1 | player_id),
  data = clean_player_df, 
  REML = TRUE
)

cat("Final model fitted successfully\n")
cat("Observations:", nrow(clean_player_df), "\n")
cat("Players:", length(unique(clean_player_df$player_id)), "\n")
cat("Clubs:", length(unique(clean_player_df$club_id)), "\n")
```

## 1. Fixed Effects Estimates

### 1.1 Complete Coefficient Table

```{r fixed_effects_table}
# Extracting fixed effects with confidence intervals
fixed_df <- broom.mixed::tidy(m_final, effects = "fixed", conf.int = TRUE) %>%
  mutate(
    # Calculating percentage change from log-coefficients
    pct_change = (exp(estimate) - 1) * 100,
    pct_lower = (exp(conf.low) - 1) * 100,
    pct_upper = (exp(conf.high) - 1) * 100,
    
    # Adding significance stars
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.10 ~ ".",
      TRUE ~ ""
    )
  )

# Displaying table
fixed_df %>%
  select(term, estimate, std.error, statistic, p.value, 
         pct_change, sig) %>%
  kable(digits = c(0, 4, 4, 2, 4, 2, 0),
        col.names = c("Predictor", "β (log)", "SE", "t", "p-value",
                      "% Change", ""),
        caption = "Fixed Effects Estimates (N = 7,023 observations, 2,215 players, 21 clubs)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE) %>%
  footnote(general = "Signif. codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1")
```

### 1.2 Key Effect Interpretations

```{r interpret_effects}
cat("\n=== INTERPRETATION OF FIXED EFFECTS ===\n\n")

# Extracting specific coefficients
beta_uefa <- fixed_df$estimate[fixed_df$term == "uefa_coef_cen"]
pct_uefa <- fixed_df$pct_change[fixed_df$term == "uefa_coef_cen"]
p_uefa <- fixed_df$p.value[fixed_df$term == "uefa_coef_cen"]

beta_goals <- fixed_df$pct_change[fixed_df$term == "goals"]
beta_assists <- fixed_df$pct_change[fixed_df$term == "assists"]
beta_yellow <- fixed_df$pct_change[fixed_df$term == "yellow_cards"]
beta_season <- fixed_df$pct_change[fixed_df$term == "season_year"]
beta_age <- fixed_df$estimate[fixed_df$term == "age_cen"]

cat("**LEVEL 3: CLUB PRESTIGE (UEFA Coefficient)**\n")
cat("================================================\n")
cat("Coefficient:", round(beta_uefa, 5), "\n")
cat("p-value:", format.pval(p_uefa, digits = 4), "\n")
cat("\nInterpretation:\n")
cat("→ Each 1-point increase in UEFA coefficient is associated with a\n")
cat("  ", round(pct_uefa, 3), "% change in player market value.\n\n")

# Calculating effect for meaningful differences
uefa_sd <- sd(clean_player_df$uefa_coefficient, na.rm = TRUE)
cat("→ A 1 SD increase in UEFA coefficient (", round(uefa_sd, 2), " points)\n")
cat("   is associated with a", round(pct_uefa * uefa_sd, 2), "% change in market value.\n\n")

cat("→ The difference between elite clubs (Real Madrid, 278 pts) and\n")
cat("   lower-ranked clubs (Villarreal, 138 pts) implies a\n")
cat("   ", round(pct_uefa * (278 - 138), 2), "% difference in market values.\n\n")

cat("\n**LEVEL 1: PERFORMANCE METRICS**\n")
cat("================================\n")
cat("Goals:    Each additional goal → +", round(beta_goals, 2), "% market value\n")
cat("Assists:  Each additional assist → +", round(beta_assists, 2), "% market value\n")
cat("Yellow Cards: Each card → ", round(beta_yellow, 2), "% change\n")
cat("Time Trend: Annual change → ", round(beta_season, 2), "% per year\n\n")

cat("\n**LEVEL 2: PLAYER CHARACTERISTICS**\n")
cat("===================================\n")
cat("Age: Each year older (vs. avg 26.5) → ", 
    round((exp(beta_age) - 1) * 100, 2), "% change\n")
```

## 2. Variance Components and ICC

```{r variance_components}
cat("\n=== VARIANCE DECOMPOSITION ===\n\n")

# Extracting variance components
vc <- as.data.frame(VarCorr(m_final))

cat("Variance Components:\n")
cat("--------------------\n")
cat("Club (τ²_club)    :", round(vc$vcov[vc$grp == "club_id"], 5), "\n")
cat("Player (τ²_player):", round(vc$vcov[vc$grp == "player_id"], 5), "\n")
cat("Residual (σ²_e)   :", round(vc$vcov[vc$grp == "Residual"], 5), "\n\n")

# Calculating ICC
total_var <- sum(vc$vcov)
icc_club <- vc$vcov[vc$grp == "club_id"] / total_var
icc_player <- vc$vcov[vc$grp == "player_id"] / total_var
icc_residual <- vc$vcov[vc$grp == "Residual"] / total_var

cat("Intraclass Correlation Coefficients:\n")
cat("-------------------------------------\n")
cat("ICC (Club)    :", round(icc_club, 4), 
    paste0("(", round(100*icc_club, 2), "% of total variance)\n"))
cat("ICC (Player)  :", round(icc_player, 4), 
    paste0("(", round(100*icc_player, 2), "% of total variance)\n"))
cat("ICC (Residual):", round(icc_residual, 4), 
    paste0("(", round(100*icc_residual, 2), "% of total variance)\n\n"))

cat("**Key Insight:**\n")
cat("After controlling for UEFA coefficient and individual characteristics,\n")
cat(round(100*icc_club, 2), "% of variance remains at the club level, suggesting\n")
cat("there are other club-level factors beyond prestige that affect valuations.\n")
```

### 2.1 Impact of UEFA on Club Variance

```{r uefa_impact}
# Comparing with model without UEFA
m_without_uefa <- lmer(
  log(usd_revenue_2024) ~ 
    season_year + yellow_cards + goals + assists +
    position + age_cen + height_cen + 
    Region + citizen_of_club_country +
    (1 | club_id) + (1 | player_id),
  data = clean_player_df, 
  REML = TRUE
)

vc_without <- as.data.frame(VarCorr(m_without_uefa))
club_var_without <- vc_without$vcov[vc_without$grp == "club_id"]
club_var_with <- vc$vcov[vc$grp == "club_id"]

var_reduction <- (club_var_without - club_var_with) / club_var_without * 100

cat("\n=== UEFA COEFFICIENT IMPACT ON CLUB VARIANCE ===\n\n")
cat("Club variance WITHOUT UEFA:", round(club_var_without, 5), "\n")
cat("Club variance WITH UEFA   :", round(club_var_with, 5), "\n")
cat("\nVariance Reduction:", round(var_reduction, 2), "%\n\n")

cat("**Interpretation:**\n")
cat("UEFA coefficient explains", round(var_reduction, 1), 
    "% of between-club differences in\n")
cat("player market values. This represents a SUBSTANTIAL portion of club-level\n")
cat("variance, indicating that club prestige/success is a major determinant.\n")
```

## 3. Model Fit Statistics

```{r model_fit}
cat("\n=== MODEL FIT STATISTICS ===\n\n")

# R-squared measures
r2_vals <- MuMIn::r.squaredGLMM(m_final)

cat("Marginal $R²$ (fixed effects only):", round(r2_vals[1], 4), "\n")
cat("  → Fixed predictors explain", 
    round(r2_vals[1] * 100, 2), "% of variance\n\n")

cat("Conditional $R²$ (fixed + random):", round(r2_vals[2], 4), "\n")
cat("  → Full model explains",
    round(r2_vals[2] * 100, 2), "% of variance\n\n")

cat("Difference:", round((r2_vals[2] - r2_vals[1]) * 100, 2), "%\n")
cat("  → Random effects add", 
    round((r2_vals[2] - r2_vals[1]) * 100, 2), 
    "% explanatory power beyond fixed effects\n")

# AIC/BIC
cat("\n\nModel Information Criteria:\n")
cat("AIC:", round(AIC(m_final), 2), "\n")
cat("BIC:", round(BIC(m_final), 2), "\n")
```

## 4. Position-Specific Effects

```{r position_effects}
cat("\n=== POSITION EFFECTS ===\n\n")

# Extracting position coefficients
pos_coefs <- fixed_df %>%
  filter(grepl("position", term)) %>%
  select(term, estimate, pct_change, p.value, sig) %>%
  mutate(
    Position = gsub("position", "", term),
    Position = factor(Position, levels = c("Attack", "Midfield", "Defender"))
  ) %>%
  arrange(Position)

cat("Position Effects (Reference: Goalkeeper)\n")
cat("-----------------------------------------\n")
print(pos_coefs %>% select(Position, pct_change, p.value, sig), row.names = FALSE)

cat("\n")
if(all(pos_coefs$p.value < 0.001)) {
  cat("All positions are significantly different from goalkeepers (p < 0.001)\n")
}

# Estimated marginal means
emm_position <- emmeans(m_final, ~ position)
pairs_position <- pairs(emm_position, adjust = "bonferroni")

cat("\n\nPairwise Comparisons (Bonferroni-adjusted):\n")
cat("-------------------------------------------\n")
print(summary(pairs_position))
```

### 4.1 Visualize Position Effects

```{r position_viz, fig.width=10, fig.height=6}
# Getting factor levels from data
pos_levels <- levels(clean_player_df$position)
region_levels <- levels(clean_player_df$Region)
citizen_levels <- levels(clean_player_df$citizen_of_club_country)

# Creating typical player profiles by position
pred_data <- data.frame(
  season_year = 2024,
  position = factor(c("Attack", "Midfield", "Defender", "Goalkeeper"),
                    levels = pos_levels),
  age_cen = 0,
  height_cen = 0,
  yellow_cards = 2,
  goals = c(15, 5, 0, 0),  # Position-appropriate goals
  assists = c(8, 5, 2, 0),  # Position-appropriate assists
  Region = factor(rep("Europe & Central Asia", 4), levels = region_levels),
  citizen_of_club_country = factor(rep("Yes", 4), levels = citizen_levels),
  uefa_coef_cen = 0  # Average UEFA coefficient
)

# Generating predictions
pred_data$log_pred <- predict(m_final, newdata = pred_data, re.form = NA)
pred_data$market_value <- exp(pred_data$log_pred)

# Plot
ggplot(pred_data, aes(x = position, y = market_value, fill = position)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::dollar(market_value, accuracy = 1)), 
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = scales::dollar_format(), 
                     expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("#E69F00", "#009E73", "#D55E00", "#56B4E9"),
                    name = "Position") +
  labs(
    title = "Predicted Market Value by Position",
    subtitle = "For typical player (avg age/height, European citizen, avg UEFA club, 2 yellow cards)\nWith position-appropriate performance: Attack (15G, 8A), Midfield (5G, 5A), Defender (0G, 2A), GK (0G, 0A)",
    x = "Position",
    y = "Predicted Market Value (2024 USD)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    legend.position = "none",
    axis.text.x = element_text(size = 12, face = "bold")
  )
```

## 5. Regional Effects

```{r regional_effects}
cat("\n=== REGIONAL EFFECTS ===\n\n")

# Extracting region coefficients
region_coefs <- fixed_df %>%
  filter(grepl("Region", term)) %>%
  select(term, estimate, pct_change, p.value, sig) %>%
  mutate(Region = gsub("Region", "", term)) %>%
  arrange(desc(pct_change))

cat("Regional Effects (Reference: East Asia & Pacific)\n")
cat("--------------------------------------------------\n")
print(region_coefs %>% select(Region, pct_change, p.value, sig), row.names = FALSE)

# EMMs for visualization
emm_region <- emmeans(m_final, ~ Region)
emm_region_df <- as.data.frame(emm_region) %>%
  mutate(
    market_value = exp(emmean),
    lower_ci = exp(asymp.LCL),
    upper_ci = exp(asymp.UCL)
  ) %>%
  arrange(desc(market_value))

cat("\n\nEstimated Market Values by Birth Region:\n")
cat("(All else equal, at average values)\n")
cat("----------------------------------------\n")
print(emm_region_df %>% 
        select(Region, market_value, lower_ci, upper_ci) %>%
        mutate(across(where(is.numeric), ~scales::dollar(., accuracy = 1))),
      row.names = FALSE)
```

## 6. UEFA Coefficient Effect Visualization

```{r uefa_viz, fig.width=10, fig.height=7}
# Getting factor levels
region_levels <- levels(clean_player_df$Region)
citizen_levels <- levels(clean_player_df$citizen_of_club_country)
pos_levels <- levels(clean_player_df$position)

# Creating prediction grid across UEFA range
uefa_range <- seq(min(clean_player_df$uefa_coefficient, na.rm = TRUE),
                  max(clean_player_df$uefa_coefficient, na.rm = TRUE),
                  length.out = 50)

# Centering based on original mean
uefa_mean <- mean(clean_player_df$uefa_coefficient, na.rm = TRUE)

pred_uefa <- data.frame(
  season_year = 2024,
  position = factor(rep("Midfield", 50), levels = pos_levels),
  age_cen = 0,
  height_cen = 0,
  yellow_cards = 2,
  goals = 10,
  assists = 5,
  Region = factor(rep("Europe & Central Asia", 50), levels = region_levels),
  citizen_of_club_country = factor(rep("Yes", 50), levels = citizen_levels),
  uefa_coefficient = uefa_range,
  uefa_coef_cen = uefa_range - uefa_mean
)

# Predictions
pred_uefa$log_pred <- predict(m_final, newdata = pred_uefa, re.form = NA)
pred_uefa$market_value <- exp(pred_uefa$log_pred)

# Adding actual club data points
club_means <- clean_player_df %>%
  group_by(Club, uefa_coefficient, uefa_rank) %>%
  summarise(
    mean_market_value = mean(usd_revenue_2024, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(pred_uefa, aes(x = uefa_coefficient, y = market_value)) +
  geom_line(color = "dodgerblue", linewidth = 2) +
  geom_point(data = club_means, 
             aes(x = uefa_coefficient, y = mean_market_value),
             size = 3, alpha = 0.6, color = "darkred") +
  geom_text(data = club_means %>% filter(uefa_rank %in% c(1, 2, 3, 20, 21)),
            aes(x = uefa_coefficient, y = mean_market_value, label = Club),
            vjust = -1, size = 3.5, fontface = "bold", color = "darkred") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Club Prestige Effect on Player Market Values",
    subtitle = "Predicted values for typical midfielder (10 goals, 5 assists)\nRed points = actual club means",
    x = "UEFA 10-Year Coefficient (2015-2025)",
    y = "Predicted Market Value (2024 USD)",
    caption = "Controlling for individual performance, age, height, position, and region"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold")
  )
```

**Interpretation:**

The blue line shows the predicted relationship between UEFA coefficient and market value, holding all else constant. Red points show actual club average market values, demonstrating good model fit.

## 7. Hypothesis Testing

### H1: Performance Effects

```{r h1_performance}
cat("\n=== HYPOTHESIS 1: Performance Metrics ===\n\n")
cat("H1: Player performance metrics (goals, assists) have positive\n")
cat("    associations with market value.\n\n")

goals_p <- fixed_df$p.value[fixed_df$term == "goals"]
assists_p <- fixed_df$p.value[fixed_df$term == "assists"]
goals_coef <- fixed_df$pct_change[fixed_df$term == "goals"]
assists_coef <- fixed_df$pct_change[fixed_df$term == "assists"]

cat("RESULT:\n")
if(goals_p < 0.001 & assists_p < 0.001) {
  cat("✓ SUPPORTED - Both goals and assists are highly significant (p < 0.001)\n\n")
  cat("  Goals: Each goal → +", round(goals_coef, 2), "% market value\n")
  cat("  Assists: Each assist → +", round(assists_coef, 2), "% market value\n\n")
  cat("Conclusion: Performance metrics strongly predict market values,\n")
  cat("supporting the 'Moneyball' hypothesis that measurable performance matters.\n")
} else {
  cat("✗ NOT SUPPORTED\n")
}
```

### H2: Club Prestige Effect

```{r h2_prestige}
cat("\n\n=== HYPOTHESIS 2: Club Prestige (UEFA) ===\n\n")
cat("H2: Club prestige (UEFA coefficient) predicts player market values\n")
cat("    beyond individual characteristics.\n\n")

uefa_p <- fixed_df$p.value[fixed_df$term == "uefa_coef_cen"]
uefa_coef <- fixed_df$pct_change[fixed_df$term == "uefa_coef_cen"]

cat("RESULT:\n")
if(uefa_p < 0.001) {
  cat("✓ STRONGLY SUPPORTED (p < 0.001)\n\n")
  cat("  UEFA effect:", round(uefa_coef, 4), "% per point\n")
  cat("  Variance explained:", round(var_reduction, 1), "% of club-level variance\n\n")
  cat("Conclusion: Club prestige has a significant 'halo effect' on player\n")
  cat("valuations. Players at elite clubs (high UEFA coefficient) command\n")
  cat("substantial premiums even after controlling for individual performance.\n")
} else {
  cat("✗ NOT SUPPORTED\n")
}
```

### H3: Age Effect

```{r h3_age}
cat("\n\n=== HYPOTHESIS 3: Age Effect ===\n\n")
cat("H3: Market value exhibits relationship with age.\n\n")

age_p <- fixed_df$p.value[fixed_df$term == "age_cen"]
age_coef <- fixed_df$pct_change[fixed_df$term == "age_cen"]

cat("RESULT:\n")
if(age_p < 0.05) {
  cat("✓ SUPPORTED (p < 0.05)\n\n")
  cat("  Age effect:", round(age_coef, 2), "% per year (relative to avg 26.5)\n\n")
  
  if(age_coef < 0) {
    cat("Interpretation: Older players have lower market values, consistent with\n")
    cat("declining physical capabilities and shorter remaining career length.\n")
  } else {
    cat("Interpretation: Older players have higher market values, possibly due to\n")
    cat("experience and peak performance years.\n")
  }
} else {
  cat("~ MARGINAL or NOT SUPPORTED\n")
}
```

## 8. Practical Implications

```{r practical_implications}
cat("\n=== PRACTICAL IMPLICATIONS ===\n\n")

cat("1. CLUB TRANSFER STRATEGY:\n")
cat("   - Players at elite clubs command", round(var_reduction, 0), 
    "% premium due to club prestige\n")
cat("   - Buying from lower-ranked clubs may offer value opportunities\n")
cat("   - Selling to elite clubs can maximize transfer fees\n\n")

cat("2. PERFORMANCE METRICS MATTER:\n")
cat("   - Goals (+", round(goals_coef, 1), "% each) more valuable than assists (+", 
    round(assists_coef, 1), "% each)\n")
cat("   - Focus on measurable output to increase player valuations\n\n")

cat("3. POSITION-SPECIFIC VALUATIONS:\n")
pos_premium <- pos_coefs$pct_change[pos_coefs$Position == "Attack"]
cat("   - Attackers valued ~", round(pos_premium, 0), 
    "% higher than goalkeepers\n")
cat("   - Position premiums reflect market demand and scarcity\n\n")

cat("4. CLUB REPUTATION INVESTMENT:\n")
cat("   - UEFA coefficient (success in European competitions) directly affects\n")
cat("     the market value of ALL players in the squad\n")
cat("   - Long-term investment in competitive success pays dividends through\n")
cat("     increased player valuations\n")
```

## 9. Model Limitations

```{r limitations}
cat("\n=== MODEL LIMITATIONS ===\n\n")

cat("1. SELECTION BIAS:\n")
cat("   - Sample includes only top 21 European clubs\n")
cat("   - Results may not generalize to lower-tier clubs\n\n")

cat("2. MISSING VARIABLES:\n")
cat("   - Injury history not included\n")
cat("   - Marketing/social media presence not captured\n")
cat("   - Contract length and release clauses unavailable\n\n")

cat("3. ENDOGENEITY:\n")
cat("   - Better players → join elite clubs (reverse causality)\n")
cat("   - UEFA coefficient partly reflects player quality\n\n")

cat("4. TIME-VARYING EFFECTS:\n")
cat("   - Model assumes constant effects over 12 years\n")
cat("   - Market dynamics may have changed post-2020 (COVID)\n\n")

cat("5. TRANSFER MARKET EFFICIENCY:\n")
cat("   - Assumes Transfermarkt valuations reflect true market prices\n")
cat("   - Actual transfer fees may differ due to negotiation, urgency, etc.\n")
```

## 10. Summary of Findings

```{r summary_table}
findings <- data.frame(
  Hypothesis = c(
    "H1: Performance Effects",
    "H2: Club Prestige (UEFA)",
    "H3: Age Effect",
    "Position Hierarchy",
    "Regional Differences",
    "Time Trend"
  ),
  
  Result = c(
    "✓ Supported",
    "✓ Strongly Supported",
    "✓ Supported",
    "✓ Significant",
    "✓ Significant",
    "✓ Significant"
  ),
  
  Key_Finding = c(
    paste0("Goals: +", round(goals_coef, 1), "%, Assists: +", round(assists_coef, 1), "%"),
    paste0(round(var_reduction, 1), "% of club variance explained, p < 0.001"),
    paste0(round(age_coef, 1), "% per year relative to mean age"),
    paste0("Attackers: +", round(pos_premium, 0), "% vs. Goalkeepers"),
    "Europe & Central Asia highest valued",
    paste0(ifelse(beta_season > 0, "+", ""), round(beta_season, 2), "% per year")
  ),
  
  stringsAsFactors = FALSE
)

kable(findings,
      col.names = c("Hypothesis/Factor", "Result", "Key Finding"),
      caption = "Summary of Main Findings") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = TRUE) %>%
  column_spec(2, bold = TRUE)
```

## Conclusions

**Main Takeaways:**

1. **Club Prestige Matters:** UEFA coefficient explains **`r round(var_reduction, 1)`%** of between-club variance in player market values, demonstrating a substantial "halo effect" where players at elite clubs command premiums independent of individual performance.

2. **Performance is Valued:** Goals and assists have significant positive effects on market value, supporting the hypothesis that measurable output drives valuations.

3. **Multilevel Structure is Essential:** The conditional $R²$ (`r round(r2_vals[2], 3)`) is much higher than marginal $R²$ (`r round(r2_vals[1], 3)`), indicating that accounting for player and club random effects is crucial for accurate modeling.

4. **Practical Value:** The model provides actionable insights for transfer strategies, player development priorities, and long-term club investment in competitive success.
