---
title: "Descriptive Statistics and Sample Characteristics"
author: "Moneyballers"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: html
theme: lumen
toc: true
execute:
  warning: false
  message: false
---

```{r setup}
library(arrow)
library(dplyr)
library(gt)
library(tidyr)
library(ggplot2)

# Reading data
repo_path <- "/Users/namomac/Desktop/Moneyball_FC/"
player_df <- read_parquet(paste0(repo_path, "inputs/player_df_with_uefa.parquet"))
```

## Sample Overview

```{r sample_overview}
cat("=== DATASET SUMMARY ===\n\n")
cat("Total Observations:", nrow(player_df), "\n")
cat("Unique Players:", length(unique(player_df$player_id)), "\n")
cat("Unique Clubs:", length(unique(player_df$Club)), "\n")
cat("Time Period:", min(player_df$season_year), "-", max(player_df$season_year), "\n")
cat("Number of Seasons:", length(unique(player_df$season_year)), "\n")
```

## Table 1: Sample Characteristics by Level

### Panel A: Level 1 - Time-Varying Measures

```{r table1a}
# Level 1 descriptives
level1_vars <- player_df %>%
  select(usd_revenue_2024, season_year, yellow_cards, red_cards, 
         goals, assists, minutes_played, Age) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable) %>%
  summarise(
    N = sum(!is.na(Value)),
    Mean = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),
    Min = min(Value, na.rm = TRUE),
    Q25 = quantile(Value, 0.25, na.rm = TRUE),
    Median = median(Value, na.rm = TRUE),
    Q75 = quantile(Value, 0.75, na.rm = TRUE),
    Max = max(Value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "usd_revenue_2024" ~ "Market Value (USD 2024)",
      Variable == "season_year" ~ "Season Year",
      Variable == "yellow_cards" ~ "Yellow Cards",
      Variable == "red_cards" ~ "Red Cards",
      Variable == "goals" ~ "Goals Scored",
      Variable == "assists" ~ "Assists",
      Variable == "minutes_played" ~ "Minutes Played",
      Variable == "Age" ~ "Player Age",
      TRUE ~ Variable
    )
  )

level1_vars %>%
  gt() %>%
  tab_header(
    title = "Level 1: Time-Varying Predictors (Measurement Occasions)",
    subtitle = paste0("N = ", nrow(player_df), " observations")
  ) %>%
  fmt_number(columns = c(Mean, SD, Q25, Median, Q75), decimals = 2) %>%
  fmt_number(columns = c(Min, Max), decimals = 0) %>%
  fmt_number(columns = N, decimals = 0, use_seps = TRUE) %>%
  cols_label(
    Variable = "Variable",
    N = "N",
    Mean = "Mean",
    SD = "SD",
    Min = "Min",
    Q25 = "Q1",
    Median = "Median",
    Q75 = "Q3",
    Max = "Max"
  )
```

### Panel B: Level 2 - Player Characteristics

```{r table1b}
# Creating player-level dataset
player_level <- player_df %>%
  group_by(player_id) %>%
  slice(1) %>%
  ungroup()

# Continuous variables
continuous_l2 <- player_level %>%
  select(Age, height_in_cm) %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(
    N = sum(!is.na(value)),
    Mean = mean(value, na.rm = TRUE),
    SD = sd(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Median = median(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = case_when(
      name == "Age" ~ "Age (years)",
      name == "height_in_cm" ~ "Height (cm)",
      TRUE ~ name
    )
  ) %>%
  select(Variable, N, Mean, SD, Min, Median, Max)

continuous_l2 %>%
  gt() %>%
  tab_header(
    title = "Level 2: Player-Invariant Characteristics (Continuous)",
    subtitle = paste0("N = ", nrow(player_level), " players")
  ) %>%
  fmt_number(columns = c(Mean, SD, Median), decimals = 2) %>%
  fmt_number(columns = c(Min, Max), decimals = 1) %>%
  fmt_number(columns = N, decimals = 0, use_seps = TRUE)

# Categorical variables
cat("\n\n")
cat("Categorical Variables:\n")
cat("---------------------\n\n")

cat("Position:\n")
print(table(player_level$position))
cat("\n")

cat("Birth Region:\n")
print(table(player_level$Region))
cat("\n")

cat("Citizen of Club Country:\n")
print(table(player_level$citizen_of_club_country))
```

### Panel C: Level 3 - Club Characteristics

```{r table1c}
# Club-level dataset
club_level <- player_df %>%
  group_by(Club) %>%
  summarise(
    n_observations = n(),
    n_players = n_distinct(player_id),
    avg_market_value = mean(usd_revenue_2024, na.rm = TRUE),
    uefa_coefficient = first(uefa_coefficient),
    uefa_rank = first(uefa_rank),
    .groups = "drop"
  ) %>%
  arrange(uefa_rank)

club_level %>%
  gt() %>%
  tab_header(
    title = "Level 3: Club Characteristics",
    subtitle = paste0("N = ", nrow(club_level), " elite European clubs")
  ) %>%
  fmt_number(columns = c(avg_market_value), decimals = 0, use_seps = TRUE) %>%
  fmt_number(columns = c(uefa_coefficient), decimals = 2) %>%
  fmt_number(columns = c(n_observations, n_players, uefa_rank), decimals = 0) %>%
  cols_label(
    Club = "Club",
    n_observations = "Observations",
    n_players = "Players",
    avg_market_value = "Avg Market Value (USD)",
    uefa_coefficient = "UEFA Coef.",
    uefa_rank = "UEFA Rank"
  ) %>%
  data_color(
    columns = uefa_rank,
    colors = scales::col_numeric(
      palette = c("darkgreen", "yellow", "red"),
      domain = c(1, 21)
    )
  )
```

## Table 2: Descriptive Statistics by Position

```{r table2}
by_position <- player_df %>%
  group_by(position) %>%
  summarise(
    N = n(),
    `Avg Age` = mean(Age, na.rm = TRUE),
    `Avg Height (cm)` = mean(height_in_cm, na.rm = TRUE),
    `Avg Goals` = mean(goals, na.rm = TRUE),
    `Avg Assists` = mean(assists, na.rm = TRUE),
    `Avg Yellow Cards` = mean(yellow_cards, na.rm = TRUE),
    `Avg Market Value (USD)` = mean(usd_revenue_2024, na.rm = TRUE),
    .groups = "drop"
  )

by_position %>%
  gt() %>%
  tab_header(
    title = "Descriptive Statistics by Player Position"
  ) %>%
  fmt_number(columns = c(`Avg Age`, `Avg Height (cm)`, `Avg Goals`, `Avg Assists`, 
                          `Avg Yellow Cards`), decimals = 2) %>%
  fmt_number(columns = `Avg Market Value (USD)`, decimals = 0, use_seps = TRUE) %>%
  fmt_number(columns = N, decimals = 0, use_seps = TRUE)
```

## Table 3: Sample Structure

```{r table3}
# Player mobility
player_clubs <- player_df %>%
  group_by(player_id) %>%
  summarise(n_clubs = n_distinct(Club), .groups = "drop")

mobility_summary <- player_clubs %>%
  mutate(
    Mobility = case_when(
      n_clubs == 1 ~ "1 club only",
      n_clubs == 2 ~ "2 clubs",
      n_clubs >= 3 ~ "3+ clubs"
    )
  ) %>%
  count(Mobility) %>%
  mutate(Percentage = round(100 * n / sum(n), 2))

# Observations per player
obs_per_player <- player_df %>%
  count(player_id, name = "n_obs")

# Creating summary table
structure_summary <- data.frame(
  Characteristic = c(
    "**Data Structure**",
    "Total Observations",
    "Unique Players",
    "Unique Clubs",
    "Time Points (Seasons)",
    "",
    "**Observations per Player**",
    "Mean (SD)",
    "Median [IQR]",
    "Range",
    "",
    "**Player Mobility**",
    "Players in 1 club only",
    "Players in 2 clubs",
    "Players in 3+ clubs",
    "% of players who transferred"
  ),
  
  Value = c(
    "",
    format(nrow(player_df), big.mark = ","),
    format(length(unique(player_df$player_id)), big.mark = ","),
    "21",
    "12 (2013-2024)",
    "",
    "",
    paste0(round(mean(obs_per_player$n_obs), 1), " (", 
           round(sd(obs_per_player$n_obs), 1), ")"),
    paste0(median(obs_per_player$n_obs), " [",
           quantile(obs_per_player$n_obs, 0.25), ", ",
           quantile(obs_per_player$n_obs, 0.75), "]"),
    paste0(min(obs_per_player$n_obs), " - ",
           max(obs_per_player$n_obs)),
    "",
    "",
    paste0(mobility_summary$n[mobility_summary$Mobility == "1 club only"], 
           " (", mobility_summary$Percentage[mobility_summary$Mobility == "1 club only"], "%)"),
    paste0(mobility_summary$n[mobility_summary$Mobility == "2 clubs"], 
           " (", mobility_summary$Percentage[mobility_summary$Mobility == "2 clubs"], "%)"),
    paste0(mobility_summary$n[mobility_summary$Mobility == "3+ clubs"], 
           " (", mobility_summary$Percentage[mobility_summary$Mobility == "3+ clubs"], "%)"),
    paste0(round(100 * sum(player_clubs$n_clubs > 1) / nrow(player_clubs), 2), "%")
  )
)

structure_summary %>%
  gt() %>%
  tab_header(
    title = "Sample Composition and Structure"
  ) %>%
  cols_label(
    Characteristic = "",
    Value = ""
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = Characteristic,
      rows = grepl("\\*\\*", Characteristic)
    )
  ) %>%
  fmt_markdown(columns = Characteristic)
```

## Visualizations

### Figure 1: Market Value Distribution

```{r fig1, fig.width=10, fig.height=6}
ggplot(player_df, aes(x = usd_revenue_2024)) +
  geom_histogram(bins = 50, fill = "dodgerblue", alpha = 0.7, color = "white") +
  scale_x_continuous(labels = scales::dollar_format(), 
                     trans = "log10",
                     breaks = c(1e5, 1e6, 1e7, 1e8)) +
  labs(
    title = "Distribution of Player Market Values",
    subtitle = "Log scale, 2024 USD (inflation-adjusted)",
    x = "Market Value (USD, log scale)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12)
  )
```

### Figure 2: UEFA Coefficient vs. Average Market Value

```{r fig2, fig.width=10, fig.height=7}
club_summary <- player_df %>%
  group_by(Club, uefa_coefficient, uefa_rank) %>%
  summarise(
    avg_value = mean(usd_revenue_2024, na.rm = TRUE),
    n_players = n_distinct(player_id),
    .groups = "drop"
  )

ggplot(club_summary, aes(x = uefa_coefficient, y = avg_value)) +
  geom_point(aes(size = n_players), alpha = 0.6, color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "dodgerblue", linewidth = 1.5) +
  geom_text(aes(label = Club), vjust = -1, size = 3, fontface = "bold") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_size_continuous(name = "Number of\nPlayers", range = c(3, 10)) +
  labs(
    title = "Club Prestige (UEFA Coefficient) vs. Average Player Market Value",
    subtitle = "Strong positive correlation: Elite clubs have higher-valued players",
    x = "UEFA 10-Year Coefficient (2015-2025)",
    y = "Average Market Value (USD 2024)",
    caption = "Point size represents number of players in sample"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12, face = "bold")
  )
```

### Figure 3: Position Distribution

```{r fig3, fig.width=8, fig.height=6}
position_counts <- player_df %>%
  count(position) %>%
  mutate(pct = round(100 * n / sum(n), 1))

ggplot(position_counts, aes(x = reorder(position, -n), y = n, fill = position)) +
  geom_col() +
  geom_text(aes(label = paste0(format(n, big.mark = ","), "\n(", pct, "%)")), 
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("#E69F00", "#009E73", "#D55E00", "#56B4E9")) +
  labs(
    title = "Distribution of Observations by Player Position",
    x = "Position",
    y = "Number of Observations"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none",
    axis.text.x = element_text(size = 12, face = "bold")
  )
```

### Figure 4: Goals and Assists by Position

```{r fig4, fig.width=10, fig.height=6}
player_df %>%
  select(position, goals, assists) %>%
  pivot_longer(cols = c(goals, assists), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = position, y = value, fill = metric)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = c("goals" = "#E69F00", "assists" = "#56B4E9"),
                    labels = c("Goals", "Assists")) +
  labs(
    title = "Distribution of Performance Metrics by Position",
    x = "Position",
    y = "Count per Season",
    fill = "Metric"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 12)
  )
```

## Summary

This dataset contains **`r format(nrow(player_df), big.mark=",")`** observations of **`r format(length(unique(player_df$player_id)), big.mark=",")`** players across **21** elite European football clubs from **2013-2024**.

**Key Characteristics:**

- **Cross-classified structure**: `r round(100*sum(player_clubs$n_clubs > 1)/nrow(player_clubs), 1)`% of players transferred between clubs
- **UEFA coefficient range**: `r round(min(club_summary$uefa_coefficient), 0)` to `r round(max(club_summary$uefa_coefficient), 0)` points
- **Market value range**: `r scales::dollar(min(player_df$usd_revenue_2024))` to `r scales::dollar(max(player_df$usd_revenue_2024))`
- **Average player age**: `r round(mean(player_df$Age), 1)` years
- **Position distribution**: Relatively balanced across Attack (`r round(100*sum(player_df$position=='Attack')/nrow(player_df), 1)`%), Midfield (`r round(100*sum(player_df$position=='Midfield')/nrow(player_df), 1)`%), Defender (`r round(100*sum(player_df$position=='Defender')/nrow(player_df), 1)`%), and Goalkeeper (`r round(100*sum(player_df$position=='Goalkeeper')/nrow(player_df), 1)`%)