---
title: "Statistical Model Specification"
author: "Moneyballers"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: html
theme: lumen
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
```

## Data Structure

Our data has a **cross-classified** structure because players can transfer between clubs over time. Specifically:

- **Level 1 (Measurement Occasions):** $n = 7,023$ season-year observations
- **Level 2 (Players):** $j = 2,215$ unique players
- **Level 3 (Clubs):** $k = 21$ elite European clubs

**Key Structural Note:** Because 456 players (20.6%) played for multiple clubs during 2013-2024, we model club and player as **crossed random effects** rather than nested.

## Mathematical Model Specification

### Full Three-Level Cross-Classified Model

**Level 1 (Within-Player, Time-Varying):**

$$
\log(Y_{tij}) = \beta_{0ij} + \beta_1(\text{season\_year}_{tij}) + \beta_2(\text{yellow\_cards}_{tij}) + \beta_3(\text{goals}_{tij}) + \beta_4(\text{assists}_{tij}) + e_{tij}
$$

where:

- $Y_{tij}$ = market value (2024 USD) for player $i$ in club $j$ at time $t$
- $e_{tij} \sim N(0, \sigma^2_e)$ is the within-player residual variance

**Level 2 (Player-Invariant):**

$$
\beta_{0ij} = \gamma_{00} + \sum_{p=1}^{3} \gamma_{0p}(\text{position}_{ij}^{(p)}) + \gamma_{04}(\text{age\_cen}_{ij}) + \gamma_{05}(\text{height\_cen}_{ij}) + \sum_{r=1}^{5} \gamma_{0r+5}(\text{Region}_{ij}^{(r)}) + \gamma_{011}(\text{citizen\_club}_{ij}) + u_{0i} + v_{0j}
$$

where:

- $\text{position}_{ij}^{(p)}$ are dummy variables for Attack, Midfield, Defender (reference: Goalkeeper)
- $\text{age\_cen}_{ij} = \text{Age}_{ij} - \overline{\text{Age}}$ (grand-mean centered)
- $\text{height\_cen}_{ij} = \text{Height}_{ij} - \overline{\text{Height}}$ (grand-mean centered)
- $\text{Region}_{ij}^{(r)}$ are dummy variables for player's birth region
- $\text{citizen\_club}_{ij}$ = 1 if player is citizen of club's country, 0 otherwise
- $u_{0i} \sim N(0, \tau^2_{\text{player}})$ is the player-specific random intercept
- $v_{0j} \sim N(0, \tau^2_{\text{club}})$ is the club-specific random intercept

**Level 3 (Club-Level):**

$$
v_{0j} = \delta_0 + \delta_1(\text{uefa\_coef\_cen}_j) + w_{0j}
$$

where:

- $\text{uefa\_coef\_cen}_j$ = UEFA 10-year coefficient (centered) for club $j$
- $w_{0j} \sim N(0, \tau^2_{\text{club\_residual}})$ is the residual club-level variance after accounting for UEFA prestige

**Rationale:** UEFA coefficient captures club prestige and historical success in European competitions, which may create a "halo effect" on player valuations independent of individual performance.

### Final Composite Model

Substituting Level 2 and 3 into Level 1:

$$
\begin{aligned}
\log(Y_{tij}) = & \; \gamma_{00} + \gamma_1(\text{season\_year}_{tij}) + \gamma_2(\text{yellow\_cards}_{tij}) + \gamma_3(\text{goals}_{tij}) + \gamma_4(\text{assists}_{tij}) \\
& + \sum_{p=1}^{3} \gamma_{0p}(\text{position}_{ij}^{(p)}) + \gamma_{04}(\text{age\_cen}_{ij}) + \gamma_{05}(\text{height\_cen}_{ij}) \\
& + \sum_{r=1}^{5} \gamma_{0r+5}(\text{Region}_{ij}^{(r)}) + \gamma_{011}(\text{citizen\_club}_{ij}) \\
& + \delta_1(\text{uefa\_coef\_cen}_j) + u_{0i} + v_{0j} + e_{tij}
\end{aligned}
$$

## Notation Table

```{r notation_table, echo=FALSE}
notation_df <- data.frame(
  Symbol = c("$Y_{tij}$", "$t$", "$i$", "$j$", 
             "$\\beta_0, \\beta_1, \\ldots$", "$\\gamma_{00}, \\gamma_1, \\ldots$",
             "$u_{0i}$", "$v_{0j}$", "$e_{tij}$",
             "$\\sigma^2_e$", "$\\tau^2_{\\text{player}}$", "$\\tau^2_{\\text{club}}$"),
  
  Description = c(
    "Market value (2024 USD) at time t for player i in club j",
    "Time index (season-year: 2013–2024)",
    "Player index (i = 1, ..., 2215)",
    "Club index (j = 1, ..., 21)",
    "Level-1 coefficients (time-varying effects)",
    "Fixed effects (population-level parameters)",
    "Player-specific random intercept",
    "Club-specific random intercept",
    "Within-player residual error",
    "Residual variance (Level 1)",
    "Between-player variance (Level 2)",
    "Between-club variance (Level 3)"
  ),
  
  Level = c("Outcome", "L1", "L2", "L3", "L1", "Fixed", 
            "Random (L2)", "Random (L3)", "Random (L1)",
            "Variance", "Variance", "Variance")
)

kable(notation_df, 
      col.names = c("Symbol", "Description", "Level"),
      escape = FALSE,
      align = c("l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Centering Decisions

### Grand-Mean Centering

We apply grand-mean centering to **age** and **height**:

$$
\text{age\_cen}_{ij} = \text{Age}_{ij} - \overline{\text{Age}} = \text{Age}_{ij} - 26.5
$$

$$
\text{height\_cen}_{ij} = \text{Height}_{ij} - \overline{\text{Height}} = \text{Height}_{ij} - 182.4 \text{ cm}
$$

**Rationale:**

1. **Interpretation:** Coefficients represent effects for a player of average age/height
2. **Reduces Multicollinearity:** Centering reduces correlation between main effects and potential interaction terms
3. **Meaningful Zero:** The intercept now represents log-market-value for an average player in the reference category

### Variables NOT Centered

- **season_year:** Kept on original scale to preserve chronological interpretation (year 2013 = baseline)
- **Performance metrics (goals, assists, yellow_cards):** Zero has natural meaning (no goals/cards)
- **Club-level predictors:** Only 21 clubs, so centering provides minimal benefit

## Model Assumptions

1. **Linearity:** Log-transformation of outcome addresses non-linearity (see diagnostic plots)
2. **Independence:** Residuals $e_{tij}$ independent conditional on random effects
3. **Normality of Residuals:** $e_{tij} \sim N(0, \sigma^2_e)$
4. **Normality of Random Effects:**
   - $u_{0i} \sim N(0, \tau^2_{\text{player}})$
   - $v_{0j} \sim N(0, \tau^2_{\text{club}})$
5. **Homoscedasticity:** Constant variance across levels of predictors
6. **No Perfect Multicollinearity:** VIF checks conducted (see diagnostics)

## Justification for Model Choices

### 1. Why Log-Transformation?

Original market values range from \$0 to \$200M with severe right skew. Log-transformation:

- Normalizes residuals (see QQ-plots in diagnostics)
- Stabilizes variance (homoscedasticity)
- Allows interpretation as percentage changes

### 2. Why Cross-Classified Instead of Nested?

```{r player_mobility, echo=FALSE}
# This will be populated from actual data in the next file
mobility_summary <- data.frame(
  Statistic = c("Players in 1 club only", 
                "Players in 2 clubs", 
                "Players in 3+ clubs",
                "% of players who transferred"),
  Value = c("1,759 (79.4%)", 
            "384 (17.3%)", 
            "72 (3.3%)",
            "20.6%")
)

kable(mobility_summary, col.names = c("Category", "Count (%)")) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

**Decision:** Use `(1 | club_id) + (1 | player_id)` (crossed) rather than `(1 | club_id/player_id)` (nested)

**Justification:** 456 players transferred at least once, violating strict nesting assumption.

### 3. Why No Random Slopes?

Initial models with random slopes for `season_year` failed to converge or were singular. This suggests:

- Insufficient between-club variation in time trends
- Model over-parameterization with only 21 clubs
- Fixed slope is adequate for this dataset

## Software Implementation

The model is fitted in R using:

```r
library(lme4)

m_final <- lmer(
  log(usd_revenue_2024) ~ 
    # Fixed effects
    season_year + yellow_cards + goals + assists +
    position + age_cen + height_cen + 
    Region + citizen_of_club_country +
    uefa_coef_cen +  # Club prestige (Level 3)
    # Random effects (crossed)
    (1 | club_id) + (1 | player_id),
  data = clean_player_df, 
  REML = TRUE  # Use REML for final model
)
```

**Estimation Method:** REML (Restricted Maximum Likelihood) for final parameter estimates, ML (Maximum Likelihood) for model comparison.

---

## Next Steps

1. Fit the model and check convergence
2. Assess model diagnostics (see `03_model_diagnostics.qmd`)
3. Interpret fixed effects and variance components (see `04_interpretation.qmd`)