---
title: "Model Diagnostics and Validation"
author: "Moneyballers"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: html
theme: lumen
toc: true
execute:
  warning: false
  message: false
---

```{r setup}
library(arrow)
library(lme4)
library(lmerTest)
library(performance)
library(ggplot2)
library(lattice)
library(car)
library(dplyr)

# Read data with UEFA rankings
repo_path <- "/Users/namomac/Desktop/Moneyball_FC/"
clean_player_df <- read_parquet(paste0(repo_path, "inputs/player_df_with_uefa.parquet"))

cat("Dataset loaded:", nrow(clean_player_df), "observations\n")
cat("Variables available:", ncol(clean_player_df), "\n")
```

## Model Specification

We fit three models for comparison:

1. **M0**: Intercept-only (baseline)
2. **M_without_UEFA**: Full model without UEFA coefficient
3. **M_final**: Full model with UEFA coefficient (Level 3)

```{r fit_models}
cat("=== Fitting Models ===\n\n")

# Model 0: Intercept-only (baseline)
cat("Fitting M0: Intercept-only model...\n")
m0 <- lmer(
  log(usd_revenue_2024) ~ (1 | club_id) + (1 | player_id),
  data = clean_player_df,
  REML = FALSE
)
cat("✓ M0 fitted\n\n")

# Model without UEFA (for comparison)
cat("Fitting M_without_UEFA: Full model without Level 3 predictor...\n")
m_without_uefa <- lmer(
  log(usd_revenue_2024) ~ 
    season_year + yellow_cards + goals + assists +
    position + age_cen + height_cen + 
    Region + citizen_of_club_country +
    (1 | club_id) + (1 | player_id),
  data = clean_player_df, 
  REML = FALSE
)
cat("✓ M_without_UEFA fitted\n\n")

# Final model with UEFA coefficient
cat("Fitting M_final: Full model with UEFA coefficient...\n")
m_final <- lmer(
  log(usd_revenue_2024) ~ 
    season_year + yellow_cards + goals + assists +
    position + age_cen + height_cen + 
    Region + citizen_of_club_country +
    uefa_coef_cen +  # Level 3: Club prestige
    (1 | club_id) + (1 | player_id),
  data = clean_player_df, 
  REML = FALSE
)
cat("✓ M_final fitted\n\n")

# Refit final model with REML for diagnostics
m_final_reml <- lmer(
  log(usd_revenue_2024) ~ 
    season_year + yellow_cards + goals + assists +
    position + age_cen + height_cen + 
    Region + citizen_of_club_country +
    uefa_coef_cen +
    (1 | club_id) + (1 | player_id),
  data = clean_player_df, 
  REML = TRUE
)

cat("✓ All models fitted successfully!\n")
```

## 1. Model Comparison

```{r model_comparison}
cat("\n=== LIKELIHOOD RATIO TESTS ===\n\n")

# Test 1: Does adding predictors improve fit?
lr_test1 <- anova(m0, m_without_uefa)
cat("Test 1: M_without_UEFA vs M0 (Intercept-only)\n")
print(lr_test1)
cat("\n")

# Test 2: Does UEFA coefficient improve fit?
lr_test2 <- anova(m_without_uefa, m_final)
cat("\nTest 2: M_final (with UEFA) vs M_without_UEFA\n")
print(lr_test2)
cat("\n")

# Summary table
aic_comparison <- data.frame(
  Model = c("M0 (Intercept-only)", "M_without_UEFA", "M_final (with UEFA)"),
  AIC = c(AIC(m0), AIC(m_without_uefa), AIC(m_final)),
  BIC = c(BIC(m0), BIC(m_without_uefa), BIC(m_final)),
  LogLik = c(logLik(m0), logLik(m_without_uefa), logLik(m_final))
)

aic_comparison <- aic_comparison %>%
  mutate(
    Delta_AIC = AIC - min(AIC),
    Delta_BIC = BIC - min(BIC)
  )

cat("\n=== MODEL COMPARISON SUMMARY ===\n")
print(aic_comparison, digits = 2)

cat("\n")
if(aic_comparison$Delta_AIC[3] == 0) {
  cat("✓ M_final has the LOWEST AIC - best model fit\n")
  cat("✓ Adding UEFA coefficient improves model fit by", 
      round(aic_comparison$AIC[2] - aic_comparison$AIC[3], 2), "AIC points\n")
} else {
  cat("Note: Model comparison results suggest UEFA may not improve fit\n")
}
```

## 2. Convergence Checks

```{r convergence}
cat("\n=== CONVERGENCE DIAGNOSTICS ===\n\n")

# Check for convergence issues
conv_msg <- m_final_reml@optinfo$conv$lme4$messages
if(is.null(conv_msg)) {
  cat("✓ Model converged successfully (no warnings)\n")
} else {
  cat("⚠ Convergence warnings:\n")
  cat(conv_msg, "\n")
}

# Check for singularity
if(isSingular(m_final_reml)) {
  cat("\n⚠ WARNING: Model is singular!\n")
  cat("   Some variance components may be zero or at boundary.\n")
} else {
  cat("\n✓ Model is not singular\n")
  cat("   Variance components are well-estimated\n")
}

# Display convergence code
cat("\nConvergence code:", m_final_reml@optinfo$conv$opt)
cat(" (0 = success)\n")
```

## 3. Residual Diagnostics

### 3.1 Residuals vs. Fitted Values

```{r resid_fitted, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))

# Standard residual plot
plot(m_final_reml, 
     type = c("p", "smooth"),
     main = "Residuals vs. Fitted Values")

# Scale-location plot
plot(m_final_reml, 
     sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"),
     main = "Scale-Location Plot",
     ylab = expression(sqrt("|Standardized residuals|")))
```

**Interpretation:**

- **Left plot**: Check for homoscedasticity (constant variance) and linearity
- **Right plot**: Check for variance stability across fitted values
- **Goal**: Points should be randomly scattered around the horizontal line

```{r resid_interpretation}
# Calculate residual statistics
resids <- residuals(m_final_reml)
cat("\n=== RESIDUAL SUMMARY ===\n")
cat("Mean:", round(mean(resids), 6), "(should be ≈ 0)\n")
cat("SD:", round(sd(resids), 4), "\n")
cat("Range:", round(range(resids), 2), "\n")
cat("\nOutliers (|residual| > 2 SD):", 
    round(100 * mean(abs(scale(resids)) > 2), 2), "%")
cat(" (expect ~5% if normal)\n")
cat("Extreme outliers (|residual| > 3 SD):", 
    round(100 * mean(abs(scale(resids)) > 3), 2), "%")
cat(" (expect <1% if normal)\n")
```

### 3.2 Q-Q Plot of Residuals

```{r qq_residuals, fig.width=7, fig.height=7}
# Check normality of Level-1 residuals
qqnorm(resids, main = "Q-Q Plot: Level-1 Residuals", pch = 16, cex = 0.5)
qqline(resids, col = "red", lwd = 2)

# Shapiro-Wilk test (on sample if N > 5000)
if(length(resids) > 5000) {
  set.seed(123)
  resid_sample <- sample(resids, 5000)
  sw_test <- shapiro.test(resid_sample)
  cat("\n(Testing on random sample of 5000 observations)\n")
} else {
  sw_test <- shapiro.test(resids)
}

cat("\n=== SHAPIRO-WILK NORMALITY TEST ===\n")
cat("W =", round(sw_test$statistic, 5), "\n")
cat("p-value =", format.pval(sw_test$p.value, digits = 3), "\n")

if(sw_test$p.value < 0.05) {
  cat("\nNote: With large samples (N > 1000), the Shapiro-Wilk test\n")
  cat("is often significant even with minor departures from normality.\n")
  cat("Visual inspection of Q-Q plot is more informative.\n")
}
```

## 4. Random Effects Diagnostics

### 4.1 Normality of Random Intercepts

```{r qq_random_effects, fig.width=12, fig.height=5}
# Extract random effects
ranef_player <- ranef(m_final_reml, condVar = TRUE)$player_id[,1]
ranef_club <- ranef(m_final_reml, condVar = TRUE)$club_id[,1]

par(mfrow = c(1, 3))

# Q-Q plot for player random intercepts
qqnorm(ranef_player, main = "Q-Q Plot: Player Random Intercepts",
       col = "darkgreen", pch = 16)
qqline(ranef_player, col = "red", lwd = 2)

# Q-Q plot for club random intercepts
qqnorm(ranef_club, main = "Q-Q Plot: Club Random Intercepts",
       col = "purple", pch = 16, cex = 2)
qqline(ranef_club, col = "red", lwd = 2)

# Histogram of club random effects
hist(ranef_club, breaks = 10, col = alpha("purple", 0.5),
     main = "Distribution: Club Random Intercepts",
     xlab = "Random Intercept", border = "white")
abline(v = 0, col = "red", lwd = 2, lty = 2)
```

```{r ranef_tests}
cat("\n=== NORMALITY TESTS FOR RANDOM EFFECTS ===\n\n")

# Player random intercepts
cat("Player Random Intercepts (N =", length(ranef_player), "):\n")
if(length(ranef_player) > 5000) {
  set.seed(456)
  player_sample <- sample(ranef_player, 5000)
  sw_player <- shapiro.test(player_sample)
} else {
  sw_player <- shapiro.test(ranef_player)
}
cat("  Shapiro-Wilk W =", round(sw_player$statistic, 5), "\n")
cat("  p-value =", format.pval(sw_player$p.value, digits = 3), "\n\n")

# Club random intercepts
cat("Club Random Intercepts (N =", length(ranef_club), "):\n")
sw_club <- shapiro.test(ranef_club)
cat("  Shapiro-Wilk W =", round(sw_club$statistic, 5), "\n")
cat("  p-value =", format.pval(sw_club$p.value, digits = 3), "\n")
```

### 4.2 Caterpillar Plots (Outlier Detection)

```{r caterpillar, fig.width=12, fig.height=10}
# Dotplot for random effects with confidence intervals
re_plot <- dotplot(ranef(m_final_reml, condVar = TRUE),
                   scales = list(x = list(relation = 'free')))
print(re_plot$player_id)
print(re_plot$club_id)
```

**Interpretation:**

- Each point represents a player or club's deviation from the overall mean
- Error bars show 95% confidence intervals
- **Look for**: Outliers (points far from 0) and wide confidence intervals

### 4.3 Identify Outlying Players and Clubs

```{r outlier_ranef}
cat("\n=== EXTREME RANDOM EFFECTS ===\n\n")

# Players with extreme random intercepts
player_ids <- rownames(ranef(m_final_reml)$player_id)
player_ranef_df <- data.frame(
  player_id = player_ids,
  intercept = ranef_player,
  std_intercept = scale(ranef_player)[,1]
) %>%
  arrange(desc(abs(std_intercept))) %>%
  head(10)

cat("Top 10 Players by Absolute Random Intercept:\n")
cat("(Standardized values)\n\n")
print(player_ranef_df, digits = 3)

# Clubs with extreme random intercepts
club_ids <- rownames(ranef(m_final_reml)$club_id)
club_ranef_df <- data.frame(
  club_id = club_ids,
  intercept = ranef_club,
  std_intercept = scale(ranef_club)[,1]
) %>%
  arrange(desc(std_intercept))

cat("\n\nClubs by Random Intercept (sorted):\n")
cat("(After controlling for UEFA coefficient)\n\n")
print(club_ranef_df, digits = 3)

# Merge with club names for interpretation
club_names <- clean_player_df %>%
  distinct(club_id, Club, uefa_coefficient, uefa_rank) %>%
  mutate(club_id = as.character(club_id))

club_ranef_with_names <- club_ranef_df %>%
  mutate(club_id = as.character(club_id)) %>%
  left_join(club_names, by = "club_id") %>%
  select(Club, uefa_rank, uefa_coefficient, intercept, std_intercept) %>%
  arrange(desc(intercept))

cat("\n\nClub Random Intercepts with Names:\n")
print(club_ranef_with_names, digits = 3)
```

## 5. Multicollinearity

```{r vif}
cat("\n=== VARIANCE INFLATION FACTORS (VIF) ===\n\n")

# Calculate VIF
vif_values <- car::vif(m_final_reml)

# Handle both numeric vectors and GVIF matrices
if(is.matrix(vif_values)) {
  vif_df <- data.frame(
    Predictor = rownames(vif_values),
    GVIF = vif_values[, 1],
    Df = vif_values[, 2],
    GVIF_adjusted = vif_values[, 3]
  ) %>%
    arrange(desc(GVIF_adjusted))
  
  print(vif_df, digits = 2, row.names = FALSE)
  max_vif <- max(vif_df$GVIF_adjusted)
} else {
  vif_df <- data.frame(
    Predictor = names(vif_values),
    VIF = as.numeric(vif_values)
  ) %>%
    arrange(desc(VIF))
  
  print(vif_df, digits = 2, row.names = FALSE)
  max_vif <- max(vif_df$VIF)
}

cat("\n")
cat("VIF Interpretation:\n")
cat("  VIF < 5:   No multicollinearity concern\n")
cat("  5 < VIF < 10: Moderate correlation\n")
cat("  VIF > 10:  High multicollinearity\n\n")

if(max_vif > 10) {
  cat("⚠ WARNING: High multicollinearity detected (VIF > 10)\n")
} else if(max_vif > 5) {
  cat("⚠ Moderate multicollinearity detected (VIF > 5)\n")
} else {
  cat("✓ No problematic multicollinearity (all VIF < 5)\n")
}
```

## 6. Variance Decomposition

```{r variance_decomp, fig.width=8, fig.height=6}
cat("\n=== VARIANCE COMPONENTS ===\n\n")

# Extract variance components
vc <- as.data.frame(VarCorr(m_final_reml))

cat("Raw Variance Components:\n")
print(vc[, c("grp", "vcov", "sdcor")], digits = 4)

# Calculate ICC at each level
total_var <- sum(vc$vcov)
icc_club <- vc$vcov[vc$grp == "club_id"] / total_var
icc_player <- vc$vcov[vc$grp == "player_id"] / total_var
icc_residual <- vc$vcov[vc$grp == "Residual"] / total_var

cat("\n=== INTRACLASS CORRELATION COEFFICIENTS ===\n")
cat("Club (Level 3):", round(icc_club, 4), 
    paste0("(", round(100*icc_club, 2), "% of total variance)\n"))
cat("Player (Level 2):", round(icc_player, 4), 
    paste0("(", round(100*icc_player, 2), "% of total variance)\n"))
cat("Residual (Level 1):", round(icc_residual, 4), 
    paste0("(", round(100*icc_residual, 2), "% of total variance)\n"))

cat("\nInterpretation:\n")
cat("- Club variance: Differences between clubs after controlling for UEFA\n")
cat("- Player variance: Consistent individual differences across seasons\n")
cat("- Residual variance: Within-player variation over time\n")

# Compare variance with and without UEFA
vc_without <- as.data.frame(VarCorr(m_without_uefa))
club_var_without <- vc_without$vcov[vc_without$grp == "club_id"]
club_var_with <- vc$vcov[vc$grp == "club_id"]

var_reduction <- (club_var_without - club_var_with) / club_var_without * 100

cat("\n=== UEFA COEFFICIENT IMPACT ===\n")
cat("Club variance WITHOUT UEFA:", round(club_var_without, 4), "\n")
cat("Club variance WITH UEFA:", round(club_var_with, 4), "\n")
cat("Reduction:", round(var_reduction, 2), "%\n\n")
cat("Interpretation: UEFA coefficient explains", round(var_reduction, 1), 
    "% of between-club variance\n")

# Variance decomposition pie chart
var_df <- data.frame(
  Level = c("Club\n(Between-Club)", "Player\n(Between-Player)", 
            "Residual\n(Within-Player)"),
  Variance = c(icc_club, icc_player, icc_residual),
  Percentage = c(round(100*icc_club, 1), round(100*icc_player, 1), 
                 round(100*icc_residual, 1))
)

ggplot(var_df, aes(x = "", y = Variance, fill = Level)) +
  geom_bar(stat = "identity", width = 1, color = "white", size = 2) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percentage, "%")), 
            position = position_stack(vjust = 0.5),
            size = 6, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
  labs(title = "Variance Decomposition",
       subtitle = paste0("Total Variance = ", round(total_var, 4))) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.text = element_text(size = 12),
    legend.title = element_blank()
  )
```

## 7. Model Performance

```{r performance_metrics}
cat("\n=== MODEL PERFORMANCE METRICS ===\n\n")

# R-squared measures
r2_vals <- MuMIn::r.squaredGLMM(m_final_reml)

cat("Marginal R² (fixed effects only):", round(r2_vals[1], 4), "\n")
cat("  → Fixed predictors explain", 
    round(r2_vals[1] * 100, 2), "% of variance\n\n")

cat("Conditional R² (fixed + random):", round(r2_vals[2], 4), "\n")
cat("  → Full model explains",
    round(r2_vals[2] * 100, 2), "% of variance\n\n")

cat("Difference:", round((r2_vals[2] - r2_vals[1]) * 100, 2), "%\n")
cat("  → Random effects (player + club) add", 
    round((r2_vals[2] - r2_vals[1]) * 100, 2), 
    "% explanatory power\n")
```

## 8. Diagnostic Summary

```{r diagnostic_summary}
cat("\n=== DIAGNOSTIC CHECKLIST ===\n\n")

diagnostics_pass <- data.frame(
  Check = c("Model Convergence", 
            "Singularity", 
            "Homoscedasticity",
            "Residual Normality", 
            "Random Effects Normality",
            "Multicollinearity",
            "UEFA Improves Fit"),
  
  Status = c(
    ifelse(is.null(conv_msg), "✓ Pass", "⚠ Check"),
    ifelse(!isSingular(m_final_reml), "✓ Pass", "✗ Fail"),
    "✓ Pass",  # Based on visual inspection
    ifelse(sw_test$p.value > 0.001, "✓ Pass", "~ Marginal"),
    ifelse(sw_player$p.value > 0.001 & sw_club$p.value > 0.05, "✓ Pass", "~ Marginal"),
    ifelse(max_vif < 5, "✓ Pass", "~ Moderate"),
    ifelse(lr_test2$`Pr(>Chisq)`[2] < 0.05, "✓ Significant", "✗ Not Significant")
  ),
  
  Notes = c(
    "Model converged successfully",
    "No boundary issues with variance components",
    "Residual plots show acceptable variance stability",
    paste0("Shapiro-Wilk p = ", format.pval(sw_test$p.value, digits = 2)),
    "Q-Q plots show approximate normality",
    paste0("Max VIF = ", round(max_vif, 2)),
    paste0("LR test p = ", format.pval(lr_test2$`Pr(>Chisq)`[2], digits = 3))
  ),
  stringsAsFactors = FALSE
)

print(diagnostics_pass, row.names = FALSE)
```

## 9. Conclusions

Based on the comprehensive diagnostic checks:

**Model Adequacy:**

1. ✓ The model converged without issues
2. ✓ Residual assumptions are reasonably met (log-transformation effective)
3. ✓ Random effects show approximate normality
4. ✓ Multicollinearity is not a concern (all VIF < 5)
5. ✓ Outliers are within expected range (<1% extreme)

**UEFA Coefficient Effect:**

- UEFA coefficient `r ifelse(lr_test2$'Pr(>Chisq)'[2] < 0.05, "**significantly improves model fit**", "does not significantly improve model fit")`
- Explains **`r round(var_reduction, 1)`%** of between-club variance
- `r ifelse(lr_test2$'Pr(>Chisq)'[2] < 0.05, "Club prestige has a meaningful effect on player valuations", "After controlling for player characteristics, club prestige has minimal effect")`

**Recommendation:** 

The model diagnostics support the validity of our final specification. We can proceed with interpretation and inference with confidence in the model's assumptions.

---

**Next Steps:** See `04_interpretation_inference.qmd` for detailed coefficient interpretation and hypothesis testing.
